\chapter{Introduction}
\label{ch:introduction}

\section{Motivation}

Gliders, paragliders, and soaring birds carry no continuous source of propulsion; to remain
airborne they extract energy from \rev{air currents in the atmosphere}, primarily from
\emph{thermals}, the buoyant updrafts that form as sunlight heats the ground. A
cross-country flight thus proceeds as a repeating cycle of three phases. In the
\emph{transition} phase the flyer glides along a nearly straight path toward the next
source of lift. If none is reached at a safe altitude, a \emph{search} phase follows, during which the flyer
probes the surrounding air mass to locate lift. Once lift is found, a \emph{climb} phase
begins, in which the flyer circles within the rising air to regain altitude. The cycle
alternates long, directed displacements with episodes of slow, localized motion.

Intermittent and directionally persistent motion of this kind is \rev{one of the settings}
in which \emph{anomalous transport} arises. Let $\mathbf{r}(t)$ denote the horizontal position of a flight, measured from take-off.
Transport is anomalous when the mean-squared displacement grows as a non-trivial power of
time, $\langle |\mathbf{r}(t)|^2\rangle \sim t^{\alpha}$ with $\alpha \neq 1$, as opposed to
the linear growth ($\alpha = 1$) of ordinary Brownian diffusion. Continuous-time random
walks and L\'evy walks are the standard \flow{modelling} frameworks for such
regimes~\cite{metzler2000,zaburdaev2015}, and anomalous transport has been reported in animal
movement\rev{~\cite{viswanathan1999}}, in disordered media\rev{~\cite{bouchaud1990}}, and in turbulent
flows\rev{~\cite{solomon1993}}. \rev{These two frameworks sit at opposite ends of a spectrum: a
plain continuous-time random walk carries no memory of direction between successive steps,
whereas a L\'evy walk is a \emph{coupled} continuous-time random walk, in which each waiting
time is tied to a ballistic displacement. The soaring cycle---directed glides interrupted by
localized climbs---falls between them, and is more naturally described by a persistent random
walk endowed with a waiting-time distribution between successive legs, so that both the leg
statistics and the correlation between successive flight directions are retained. Isolating
such a minimal model is one aim of the analysis developed later.}
Whether the trajectories of real soaring flights belong to an anomalous regime, and of which
type, is the question addressed in this thesis.

Soaring has previously been studied mainly as a control problem---how to centre and exploit
a thermal efficiently, including by reinforcement learning~\cite{reddy2016}. The present
work is statistical instead: it asks which transport properties emerge at the level of the
ensemble of observed trajectories. Its direct precedent is the study of Vilpellet et
al.~\cite{vilpellet2026}, who analysed a large set of human soaring flights and found a
horizontal transport that is sub-ballistic and approximately independent of aircraft type
(Hurst exponent $H \approx 0.88$, i.e.\ $\langle|\mathbf{r}(t)|^2\rangle\sim t^{2H}$,
between diffusive $H=1/2$ and ballistic $H=1$\rev{, so that a value $1/2<H<1$ marks
superdiffusive transport}). That study also introduced a
phase-resolved analysis based on the transition--search--climb cycle. This thesis builds
on it: it enlarges the dataset\rev{, revisits the cleaning and filtering that precede any
analysis,} and re-examines both the phase segmentation and the transport analysis.

\section{Objectives}

The thesis pursues \rev{the following} objectives:
\begin{enumerate}
  \item to assemble a large, reproducible, and documented dataset of real soaring flights;
  \item \rev{to clean and filter the raw GPS logs into analysis-ready trajectories, and to
        define suitable transport observables;}
  \item \rev{to define a segmentation strategy for the transition--search--climb cycle and
        characterize the motion statistics within each phase;}
  \item \rev{to test for anomalous transport and to build a minimal stochastic model able
        to reproduce the empirical statistics, rather than only comparing the data against
        existing models;}
  \item \rev{to extend the dataset to competition flights and examine how flying in
        competition changes the motion relative to solo flights.}
\end{enumerate}
A first dataset---the archive of the \emph{Coupe F\'ed\'erale de Distance} (CFD) of the
French free-flight federation (FFVL), covering paraglider and hang-glider flights---has
been assembled and verified and is described here; \rev{data collection continues and will
be extended to further sources, in particular sailplanes and competition flights}
(Sec.~\ref{sec:acquisition}). \rev{The dataset and its cleaning-and-filtering pipeline are
described here; the segmentation, the transport analysis and modelling, and the extension to
competition flights define the work ahead} (Chapter~\ref{ch:next-steps}).

\section{Scope of this document}

This document reports the current state of the project and is revised as the work proceeds.
\rev{Its statistics, tables, and figures are produced directly from the dataset by the
analysis code, so the numbers quoted in the text cannot drift from the data they describe.}
\rev{Chapter~\ref{ch:dataset} follows the data end to end: its acquisition and the
reproducibility of the pipeline (Sec.~\ref{sec:acquisition}); the tracklog format and the
flight catalog (Secs.~\ref{sec:igcformat}--\ref{sec:catalog}); the glider categories and
the coverage statistics (Secs.~\ref{sec:glider}--\ref{sec:coverage}); the known data-quality
caveats (Sec.~\ref{sec:dataquality}); the pre-processing
that turns the raw logs into analysis-ready trajectories (Sec.~\ref{sec:preproc}); a
first characterization of the cleaned dataset (Sec.~\ref{sec:prelim})}\flow{; and a
one-map summary of the whole pipeline that closes the chapter
(Sec.~\ref{sec:pipelinemap}).}
Chapter~\ref{ch:next-steps} sets out the planned transport analysis and modelling.

Throughout, the discussion keeps to the scientific and statistical reasoning behind each
choice. Implementation-level detail (code modules, configuration values, sample sizes,
computation times, reproduction commands) is collected in \nameref{app:implementation},
from p.~\pageref{app:implementation}. The appendix is organised chapter by chapter to
mirror the body; each body section points to its entry as ``implementation details:
Sec.~X''.
