\chapter{Data acquisition}
\label{ch:acquisition}

\section{Source: the FFVL Coupe F\'ed\'erale de Distance}

The data come from the \emph{Coupe F\'ed\'erale de Distance} (CFD), the cross-country
competition run by the French free-flight federation (FFVL). The federation operates two
parallel competitions on separate sites: one for paragliders
(\texttt{parapente.ffvl.fr}) and one for hang gliders (\texttt{delta.ffvl.fr}). Pilots
upload their flights; each entry carries metadata (pilot, date, take-off and landing sites,
scored distance, duration, glider) and, in most cases, the raw GPS tracklog in IGC format.
Both sites expose identical URL structure and machine-readable XML exports, so the same
acquisition pipeline handles both. A ``season'' is identified by its starting year,
e.g.\ the year~1999 denotes the 1999--2000 season; the paraglider archive starts in
1999--2000 and the hang-glider archive in 2001--2002.

\section{What is collected}

For every season the pipeline retrieves two things:
\begin{itemize}
  \item the \textbf{season XML export}, archived verbatim as provenance
        (\texttt{raw\_xml/<year>.xml}); it is the authoritative list of flights and links;
  \item the \textbf{IGC tracklog files} referenced by the XML, one file per flight that
        has a track.
\end{itemize}
From these, a derived \textbf{catalog} is built (\texttt{catalog.csv}, one row per flight,
metadata plus the local path of its track) together with a per-season index
(\texttt{seasons\_index.csv}). The raw data (tens of gigabytes) live on an external disk
and are never committed to the repository; only the small per-season index is versioned,
and it is the source of the statistics in Chapter~\ref{ch:dataset}.

\section{Pipeline and tooling}

Acquisition is performed by a dedicated, installable Python package
(\texttt{soaring.acquisition.ffvl}), exposed through two command-line tools:
\texttt{soaring-para} for paraglider data and \texttt{soaring-delta} for hang-glider data.
Both share the same subcommands: \texttt{fetch-xml} archives the season XMLs,
\texttt{download} downloads the IGC files, \texttt{build-catalog} rebuilds the catalog
and index, \texttt{status} reports per-season coverage, and \texttt{verify} checks the
integrity of the files on disk.
The design favours robustness for a long-running, large job:
\begin{description}
  \item[Resumable downloads.] Files already present are skipped, so the job can be
        interrupted and resumed freely.
  \item[Atomic writes.] Each file is written to a temporary \texttt{.part} and then
        renamed, so a file under its final name is always complete -- important on the
        exFAT external disk, which has no journaling.
  \item[Content validation.] A response is accepted only if it looks like a real IGC file
        (an \texttt{A} header record followed by at least one \texttt{B} fix record);
        HTML error pages and truncated responses are rejected and retried.
  \item[Courteous networking.] A small bounded thread pool, per-request timeouts,
        exponential backoff, and a minimum delay between requests avoid hammering the
        server; a browser TLS fingerprint is impersonated to pass the site's Cloudflare
        challenge.
\end{description}

\section{Reproducibility}

The destination directory (\texttt{data\_root}) is not hard-coded: it is set per machine
through an environment variable (\texttt{SOARING\_PARA\_DATA\_ROOT} for paragliders,
\texttt{SOARING\_DELTA\_DATA\_ROOT} for hang gliders) or a configuration file shipping a
placeholder (\texttt{configs/para\_download.yaml} and \texttt{configs/delta\_download.yaml}),
so no machine-specific path is committed. Every command validates this setting at start-up
and aborts with an explicit message if the target is missing, rather than failing obscurely.
Archiving the season XMLs verbatim keeps the full provenance: the catalog can always be
rebuilt from them with \texttt{build-catalog}.

\section{Integrity}

Because the data sit on an external drive, the \texttt{verify} subcommand re-scans every
downloaded file on disk, reusing the same IGC validation applied at download time, and
flags truncated or invalid files, leftover \texttt{.part} fragments, and read errors. The
FFVL paraglider collection has been verified in full: all \num{\StatDownloaded} downloaded
tracklogs pass validation (Chapter~\ref{ch:dataset}).

\section{Further data sources}
The FFVL CFD is the primary source of the present study, but the dataset is not assumed to
be final. The paraglider data (\textasciitilde203\,000 flights, \textasciitilde186\,000
tracklogs) are the main analysis target; the hang-glider data (\textasciitilde9\,300
flights, \textasciitilde6\,750 tracklogs) are collected in parallel and will be used for
cross-type comparison. Additional public soaring repositories---notably XContest---are
being pursued to enlarge and diversify the sample. These expose comparable flight metadata
and IGC tracks and can be acquired with the same scraping approach and merged into the same
catalog.
