\chapter{The dataset}
\label{ch:dataset}

\section{The IGC tracklog format}

A flight track is stored as an \emph{IGC file}, the plain-text format defined by the
FAI/IGC Technical Specification for IGC-approved GNSS flight recorders~\cite{fai_igc_spec};
it is the standard output of the recorders used in free-flight competitions. Its layout is
therefore fixed by that standard, not inferred from the files, and the acquisition
validates every downloaded track against it (Chapter~\ref{ch:acquisition}), so conformance
is checked across the whole archive.

Each line is a \emph{record} whose type is set by its first character -- a code assigned by
the standard, not chosen by the logger. The records relevant here are:
\begin{description}
  \item[\texttt{A}] the flight-recorder/manufacturer identifier (first line of the file);
  \item[\texttt{H}] header records (flight date, pilot, glider, geodetic datum, \dots);
  \item[\texttt{I}] the declaration of the optional extra fields appended to every fix;
  \item[\texttt{B}] the GPS fixes -- the body of the file.
\end{description}
A \texttt{B} record stores its fields at fixed character positions. Decoded against the
standard, the example
\begin{center}
\texttt{B\,094329\,4933044N\,00005193E\,A\,00095\,00149}
\end{center}
reads as: UTC time \texttt{09:43:29}; latitude $49^\circ33.044'$\,N and longitude
$000^\circ05.193'$\,E on the WGS84 datum; a fix-validity flag; and \emph{two} altitudes in
metres -- the barometric (pressure) altitude (\SI{95}{\metre}) followed by the GNSS
altitude (\SI{149}{\metre}).

The validity flag is fixed by the standard to two values and certifies the \emph{GNSS
altitude} rather than the horizontal position: \texttt{A} marks a valid three-dimensional
fix (usable GNSS altitude), whereas \texttt{V} marks a two-dimensional fix or a GNSS
drop-out, in which case the standard requires the GNSS-altitude field to be written as
zero. A \texttt{V} fix can thus still carry a usable latitude/longitude and a valid
barometric altitude.

Decoding a \texttt{B} record from fixed character positions can succeed numerically while
the result is not a valid encoding at all -- a corrupted byte can still parse as a digit.
The parser therefore checks, at decode time and independently of any flight-dynamics
judgement, that the time is a valid 24-hour \texttt{hh:mm:ss}, that the arc-minutes field of
latitude and longitude is \texttt{<\,60} (the \texttt{DDMM.mmm} encoding is undefined
otherwise), that the hemisphere letter is one of the two valid characters, and that the
decoded latitude/longitude fall within $[-90,90]$/$[-180,180]$ degrees; a record failing any
of these is dropped as unparseable, on the same footing as a record with the wrong length or
a non-numeric field; this check is implemented in the parser now and covered by unit tests.
It is a \emph{format}-validity check -- is the record even a legal encoding -- distinct from
the \emph{plausibility} checks of fix-level cleaning (Chapter~\ref{ch:next-steps}, e.g.\ the
inter-fix speed bound), which are designed and thresholded but not yet implemented: they
target fixes that decode validly but would be physically implausible given the rest of the
trajectory.

The two altitude channels differ in reference and in noise -- the
barometric value is referenced to the ICAO standard atmosphere and is smooth, the GNSS
value is referenced to the WGS84 ellipsoid and is vertically noisier -- and on older
loggers one channel may be absent (recorded as zero). The analysis adopts the
\emph{barometric} channel for the vertical dynamics, for the signal-quality reasons set out
and quantified in Chapter~\ref{ch:next-steps}; this choice matters because the vertical
velocity derived from it drives the phase segmentation.

The sequence of \texttt{B} records is thus a time-stamped three-dimensional trajectory. The
sampling period is typically a few seconds, is not guaranteed to be uniform, and varies
across recording devices -- a point the analysis must handle explicitly.

\section{Flight metadata and the catalog}

Every flight in the season XML carries a set of metadata as record attributes, which the
acquisition preserves. Beyond the identifiers and links, the fields used downstream are:
the \textbf{date}; the \textbf{pilot} and \textbf{club}; the \textbf{take-off} and
\textbf{landing} sites and the French \textbf{department}; the glider, given as both a model
name (\texttt{aile}) and a class (\texttt{aile\_class}, Section~\ref{sec:glider}); and the
competition quantities -- the \textbf{flight type} (the \emph{scoring} category: free
distance, out-and-return, flat or FAI triangle, hike-and-fly, \dots), the scored
\textbf{distance} and \textbf{points}, and, where present, the \textbf{duration} and average
\textbf{speed}. The flight type is a scoring label and must not be read as a glider or
manoeuvre type.

From these, the pipeline builds a single derived table, the \textbf{catalog}
(\texttt{catalog.csv}), with \emph{one row per flight} across all seasons: the metadata
above plus a \texttt{local\_path} linking each flight to its track on disk and a
\texttt{downloaded} flag. It is the working index of the project -- a flight can be traced
to its file, and back (through the identifier embedded in the file name) to its metadata and
public page, without any external lookup. A smaller \textbf{season index}
(\texttt{seasons\_index.csv}), versioned in the repository, summarises each season by the
number of flights declared, carrying a track, and downloaded; it is the source of the
statistics below. Both tables are regenerated from the archived XMLs on demand
(\texttt{build-catalog}).

\section{Glider categories}
\label{sec:glider}

Soaring kinematics depend strongly on the aircraft, so the glider must be identifiable for
any stratified analysis. Three broad categories exist, in increasing order of performance:
\textbf{paragliders} (flexible canopy; slowest, glide ratio $\sim\!8$--$11$, tight turns),
\textbf{hang gliders} (rigid-framed delta wing; faster, glide ratio $\sim\!10$--$15$), and
\textbf{sailplanes} (rigid aircraft; fastest, glide ratio $\gtrsim\!25$). They differ
several-fold in cruise speed and turn radius -- which is why the reference
study~\cite{vilpellet2026} reports its transport statistics separately for each.

In practice the category is set by the \emph{data source}, since the public databases are
organised by discipline. The FFVL runs the Coupe F\'ed\'erale de Distance as two parallel
competitions -- one for \emph{paragliders} and one for \emph{hang gliders} -- so the present
dataset spans \emph{both} glider types: the glider-type stratum has two levels, and the
cross-type comparison of the reference study~\cite{vilpellet2026} is therefore already
possible between paragliders and hang gliders. The third category, \emph{sailplanes}, is
collected by different portals and is not yet in hand (Chapter~\ref{ch:acquisition}); until
it is added the cross-type analysis is limited to the two FFVL disciplines.

Within each discipline the flights can be further stratified by \emph{performance class},
recorded in \texttt{aile\_class}; the two disciplines use different class systems. FFVL
labels \emph{paragliders} on the EN/AFNOR certification ladder -- EN\,A (\enquote{A ou 1},
most stable, recreational), EN\,B (\enquote{B ou 1-2}), EN\,C (\enquote{C ou 2}), EN\,D
(\enquote{D ou 2-3}), the \enquote{CIVL Competition Class} (CCC) racing wings, plus tandem
(\enquote{biplace}) and non-certified (\enquote{non homologu\'ee}) entries. \emph{Hang
gliders} follow the FAI/CIVL class system, which in this data reduces to flexible wings
(\enquote{Delta Class 1} and the intermediate \enquote{Delta Class Sport}) and rigid wings
(\enquote{Rigide Class 2} and \enquote{Rigide Class 5}). In both cases the ladder tracks
performance -- higher classes fly faster and flatter -- so it is the natural within-source
stratification variable, playing the role that glider type plays across sources.

Two caveats bear on any use of \texttt{aile\_class}. First, the field must be normalised
before it can serve as a stratum: label variants are merged, and a small fraction of rows
carry a blank or a placeholder (\enquote{0}) class, for which the class is recovered from the
wing model (\texttt{aile}) where possible. Second, the class certifies the wing, not the
pilot. Both points are handled at the pre-processing stage (Chapter~\ref{ch:next-steps}).

\section{Coverage and statistics}

The statistics below describe the FFVL CFD dataset currently in hand; further sources may be
added later (Chapter~\ref{ch:acquisition}). It comprises \num{\StatTotalFlights} flights in
total, of which \num{\StatTotalWithIgc} carry a GPS tracklog and \num{\StatTotalDownloaded}
have been downloaded and verified, split across two disciplines.

The \emph{paraglider} collection spans \StatParaNumSeasons\ seasons, from
\StatParaSeasonFirst\ to \StatParaSeasonLast: \num{\StatParaTotalFlights} flights, of which
\num{\StatParaWithIgc} (\StatParaPctWithIgc\,\%) carry a tracklog and
\num{\StatParaDownloaded} (\StatParaPctDownloaded\,\%) are downloaded and verified
(Table~\ref{tab:seasons-para}). The \emph{hang-glider} collection spans \StatHangNumSeasons\
seasons, from \StatHangSeasonFirst\ to \StatHangSeasonLast: \num{\StatHangTotalFlights}
flights, of which \num{\StatHangWithIgc} (\StatHangPctWithIgc\,\%) carry a tracklog and
\num{\StatHangDownloaded} (\StatHangPctDownloaded\,\%) are downloaded and verified
(Table~\ref{tab:seasons-hang}). The hang-glider archive is roughly twenty times smaller,
reflecting the smaller active population of the discipline.

In both disciplines the volume of flights grows by orders of magnitude over the period,
reflecting the adoption of GPS loggers and online scoring: the early seasons contribute a
handful of tracks each, while recent seasons contribute thousands.

\input{generated/seasons_table_para}
\input{generated/seasons_table_hang}

\section{Data quality and caveats}

A few limitations are already known and will shape the analysis:
\begin{itemize}
  \item \textbf{Flights without a track.} About \SI{8}{\percent} of paraglider entries and
        about \SI{27}{\percent} of hang-glider entries have no IGC file (older or manually
        declared flights); they carry metadata only.
  \item \textbf{A small residue of unrecoverable files.} A few flights advertise a track
        that the server does not actually serve as a valid IGC (broken or placeholder
        links); these are recorded as failures and excluded.
  \item \textbf{Heterogeneous logging.} Sampling rate, altitude source, and accuracy depend
        on the device; trajectories must be resampled and quality-filtered before any
        ensemble statistic is computed.
  \item \textbf{Altitude channel.} The barometric channel is adopted for the vertical
        dynamics (Chapter~\ref{ch:next-steps}), but a sizeable minority of loggers record no
        pressure altitude -- over a quarter of paraglider and roughly a sixth of hang-glider
        flights -- which fall back to the GNSS altitude, flagged as such.
  \item \textbf{Glider metadata.} Flights are paragliders or hang gliders, tagged by source
        (Section~\ref{sec:glider}); the class field \texttt{aile\_class} uses a different
        system in each discipline and must be normalised before it is used to stratify.
  \item \textbf{Incomplete dates.} Some historical entries have placeholder dates (e.g.\
        \texttt{0000-00-00}); these are kept but flagged.
\end{itemize}
These caveats motivate the pre-processing described in Chapter~\ref{ch:next-steps}.
