\chapter{Next steps}
\label{ch:next-steps}

With the dataset in place and verified, the work proceeds as follows.

\section{Trajectory pre-processing}
Each IGC file is parsed into a sequence of fixes $(t_i,\varphi_i,\lambda_i,h_i)$ -- time,
geodetic latitude and longitude, and altitude. The format records \emph{two} altitudes per
fix, a barometric (pressure) altitude and a GNSS (GPS) altitude; they differ in both
reference and noise (the barometric value is smooth but tied to a pressure setting, the GPS
value is absolute but vertically noisier). Which one to use -- or whether to combine them --
matters for the vertical velocity that drives the segmentation; this choice is left open for
now and will be made at the analysis stage rather than fixed by default. The fixes are then resampled to a common cadence to absorb device
heterogeneity, and quality-filtered (fixes with a bad validity flag, implausible speeds, or
large time gaps are dropped). The output is a tidy, analysis-ready representation produced
by a future \texttt{soaring.analysis} sub-package.

\paragraph{From geographic to local Cartesian coordinates.}
Latitudes and longitudes are angles on a curved surface and cannot be differenced directly
to obtain distances; they are mapped to a local Cartesian frame in two steps, following the
convention of Vilpellet et al. First, each fix is converted to Earth-Centred,
Earth-Fixed (ECEF) coordinates on the WGS84 ellipsoid (semi-major axis
$a=\SI{6378137}{\metre}$, flattening $f=1/298.257223563$, $e^2=f(2-f)$),
\begin{equation}
N(\varphi)=\frac{a}{\sqrt{1-e^2\sin^2\varphi}},\qquad
\begin{aligned}
X &= (N+h)\cos\varphi\cos\lambda,\\
Y &= (N+h)\cos\varphi\sin\lambda,\\
Z &= \big(N(1-e^2)+h\big)\sin\varphi,
\end{aligned}
\end{equation}
where $N(\varphi)$ is the prime-vertical radius of curvature. Second, taking the first fix
$(\varphi_0,\lambda_0,h_0)$ of the flight as the origin, the ECEF displacement
$\Delta\mathbf{X}=\mathbf{X}-\mathbf{X}_0$ is rotated into the local East--North--Up (ENU)
frame tangent to the ellipsoid at that point,
\begin{equation}
\begin{pmatrix}E\\N\\U\end{pmatrix}=
\begin{pmatrix}
-\sin\lambda_0 & \cos\lambda_0 & 0\\
-\sin\varphi_0\cos\lambda_0 & -\sin\varphi_0\sin\lambda_0 & \cos\varphi_0\\
\cos\varphi_0\cos\lambda_0 & \cos\varphi_0\sin\lambda_0 & \sin\varphi_0
\end{pmatrix}
\Delta\mathbf{X}.
\end{equation}
The ENU frame \emph{is} the local Cartesian frame: its horizontal plane $(E,N)$ is the
``north--east'' plane and $U$ is the local vertical, so there are not two distinct
coordinate systems but one, reached through the intermediate (global) ECEF representation.
Over the spatial extent of a single flight (tens of kilometres) the tangent-plane
approximation is accurate to well below the GPS noise.
% TODO: add a schematic figure (curved Earth + tangent ENU plane at the take-off point).

\paragraph{Notation.}
Throughout, $\mathbf{r}(t)=\big(x(t),y(t)\big)=\big(E(t),N(t)\big)$ denotes the
\emph{horizontal} position in this local frame and $z(t)=U(t)$ the altitude. By
construction each trajectory starts at the origin, $\mathbf{r}(0)=\mathbf{0}$, with the
clock reset to $t=0$ at take-off, so $t$ is the elapsed flight time. The horizontal
velocity is $\mathbf{v}(t)=\dot{\mathbf{r}}(t)$ and the horizontal displacement over a lag
$t$ is $\Delta\mathbf{r}(t)=\mathbf{r}(t_0+t)-\mathbf{r}(t_0)$.

\section{Observables on the whole trajectory}
\label{sec:obs-global}
A first group of observables requires \emph{no} segmentation: it can be measured directly on
the full trajectories and already probes the transport at the global level. For each, we
state what it is, how it is computed, and what it is for.

\paragraph{Transport and diffusion.}
\begin{description}
  \item[Mean-squared displacement (MSD).] With every flight starting at the origin
        ($\mathbf{r}(0)=\mathbf{0}$), the \emph{ensemble-averaged} MSD is the mean squared
        horizontal distance reached after an elapsed time $t$,
        \begin{equation}
          \mathrm{MSD}(t)=\big\langle\,|\mathbf{r}(t)|^2\big\rangle,
        \end{equation}
        where $\langle\cdot\rangle$ is an average \emph{over flights} (an ensemble average) at
        fixed $t$. Its growth law
        $\mathrm{MSD}(t)\sim t^{\alpha}$ fixes the transport regime ($\alpha=1$ Brownian,
        $\alpha>1$ super-diffusive, $\alpha<1$ sub-diffusive; equivalently the Hurst exponent
        $H=\alpha/2$). This is the primary test for anomalous transport.
  \item[Time-averaged MSD (TA-MSD).] A second MSD estimator is built entirely
        \emph{within} a single flight. For flight $k$ of duration $T_k$, one slides a window
        of width $t$ along the trajectory $\mathbf{r}^{(k)}$ and averages the squared
        displacement over all starting times $t_0$ \emph{inside that flight}:
        \begin{equation}
          \overline{\delta^2}^{(k)}(t)\;\equiv\;
          \frac{1}{T_k-t}\int_0^{T_k-t}
          \bigl|\mathbf{r}^{(k)}(t_0+t)-\mathbf{r}^{(k)}(t_0)\bigr|^2\,\mathrm{d}t_0 .
        \end{equation}
        The average over $t_0$ here is a \emph{time} average internal to flight $k$; it has
        nothing to do with the ensemble average $\langle\cdots\rangle$ over different flights
        used in the MSD. Because a single-flight curve $\overline{\delta^2}^{(k)}(t)$ is
        noisy, the operationally useful quantity is its mean over all flights of comparable
        duration $T$:
        \begin{equation}
          \mathrm{TAMSD}(t,T)\;\equiv\;
          \Bigl\langle\overline{\delta^2}^{(k)}(t)\Bigr\rangle_{T_k\approx T}.
        \end{equation}

        \smallskip
        \textbf{Ergodic case.} For an ergodic process,
        $\mathrm{TAMSD}(t,T)\to\mathrm{MSD}(t)$ as $T\to\infty$, independently of $T$, and
        the individual values $\overline{\delta^2}^{(k)}(t)$ concentrate around the common
        limit (the distribution collapses to a delta function).

        \smallskip
        \textbf{Non-ergodic case: two independent signatures.}
        \begin{enumerate}[label=(\roman*)]
          \item \emph{Aging.} $\mathrm{TAMSD}(t,T)$ differs from $\mathrm{MSD}(t)$ and
                retains a systematic dependence on $T$: flights of different duration yield
                different curves. This is a strong indicator of non-ergodicity, but it is
                sufficient, not necessary -- some non-ergodic processes show only a weak
                $T$-dependence.
          \item \emph{Non-self-averaging.} The distribution of
                $\overline{\delta^2}^{(k)}(t)$ across trajectories remains broad even as
                $T\to\infty$: individual flights give systematically different TA-MSD values
                and the distribution does not collapse. This is the more fundamental
                signature: it persists even when aging is weak.
        \end{enumerate}

        \smallskip
        We therefore monitor both $\mathrm{TAMSD}(t,T)$ and the full distribution of
        $\overline{\delta^2}^{(k)}(t)$ across trajectories at fixed $t$ and $T$. The
        canonical illustration -- a CTRW with power-law waiting times -- is worked out
        analytically in Appendix~\ref{app:ctrw}, where both EA-MSD$\neq$TA-MSD and the
        $T$-dependence are derived explicitly.
  \item[Propagator.] Since each flight starts at the origin,
        $\mathbf{r}(0)=\mathbf{0}$, the position at time $t$ and the displacement from the
        start coincide: $\mathbf{r}(t)=\mathbf{r}(t)-\mathbf{r}(0)$. We therefore define the
        propagator as the probability density of the position vector $\mathbf{x}\in\mathbb{R}^2$
        at time $t$,
        \begin{equation}
          P(\mathbf{x},t)\,\mathrm{d}^2x
          = \Pr\!\bigl(\mathbf{r}(t)\in[\mathbf{x},\,\mathbf{x}+\mathrm{d}\mathbf{x}]\bigr).
        \end{equation}
        For a \emph{self-similar} process with exponent $H$ -- including ordinary Brownian
        motion ($H=\tfrac{1}{2}$) as a special case -- the propagator obeys the scaling form
        \begin{equation}
          P(\mathbf{x},t) = t^{-2H}\,G\!\left(\frac{\mathbf{x}}{t^H}\right),
          \label{eq:prop-scaling-2d}
        \end{equation}
        where $G:\mathbb{R}^2\to\mathbb{R}_{>0}$ is a time-independent scaling function
        (normalised, $\int G\,\mathrm{d}^2u=1$). The single number $H$ controls the spread:
        $\langle|\mathbf{r}(t)|^2\rangle\sim t^{2H}$, consistently with the MSD definition.

        In practice one works with the \textbf{one-dimensional marginal} obtained by
        projecting onto a single Cartesian component $x$,
        \begin{equation}
          P(x,t) \equiv \int_{-\infty}^{+\infty} P\!\left((x,y),t\right)\mathrm{d}y .
        \end{equation}
        Assuming the process is \emph{isotropic}, $G(\mathbf{u})=G(|\mathbf{u}|)$, the
        integral can be evaluated by the substitution $y=t^H v$:
        \begin{equation}
          P(x,t) = t^{-2H}\int_{-\infty}^{+\infty}
          G\!\left(\sqrt{\!\left(\tfrac{x}{t^H}\right)^{\!2}+v^2}\right)t^H\,\mathrm{d}v
          = t^{-H}\,F\!\left(\frac{x}{t^H}\right),
          \label{eq:marginal-scaling}
        \end{equation}
        where $F(u)\equiv\int_{-\infty}^{+\infty}G(\sqrt{u^2+v^2})\,\mathrm{d}v$ inherits
        the normalisation $\int F\,\mathrm{d}u=1$ from $G$. The marginal therefore satisfies
        the same scaling form as the 2D propagator, with the same exponent $H$ but in one
        dimension. The shape of $F$ distinguishes transport regimes: a Gaussian $F$
        characterises ordinary diffusion, while heavy power-law or stretched-exponential tails
        signal anomalous transport.

        The marginal $P(x,t)$ is the most practical object to estimate from data (use either
        the East or the North component, or both for an isotropy check) and to collapse onto
        the scaling form~\eqref{eq:marginal-scaling} by rescaling the axis by $t^H$.
  \item[Probability of return to the origin.] The value of the propagator at its centre,
        $P_0(t)\equiv P(\Delta\mathbf{r}=\mathbf{0},t)$, i.e.\ the height of the peak. For a
        self-similar process it decays as a power law set by the same exponent,
        $P_0(t)\sim t^{-dH}$ in $d$ dimensions ($t^{-2H}$ in the plane, $t^{-H}$ for a single
        component). It therefore yields $H$ from the \emph{bulk} of the distribution rather
        than from its tails, which is valuable when heavy tails make the variance -- and hence
        the MSD -- ill-defined; a change of slope in $\log P_0$ versus $\log t$ exposes a
        crossover between regimes, and a peak exponent that disagrees with the MSD exponent is
        itself informative, pointing to multi-scaling. (This is the route Mantegna and Stanley
        used to read off the index of L\'evy-like walks.)
  \item[Non-Gaussianity.] How far the propagator departs from a Gaussian, measured by the
        non-Gaussian parameter, in the plane
        \begin{equation}
          \alpha_2(t)=\frac{\big\langle|\Delta\mathbf{r}(t)|^4\big\rangle}
          {2\,\big\langle|\Delta\mathbf{r}(t)|^2\big\rangle^{2}}-1 ,
        \end{equation}
        which vanishes for a Gaussian and is positive for heavier-than-Gaussian tails. It is
        evaluated on $P(\Delta\mathbf{r},t)$ \emph{at each lag} $t$: it is therefore a function
        of $t$, one number per lag, tracing a curve $\alpha_2(t)$ that pinpoints the time
        scales over which the displacement statistics are non-Gaussian (typically large at the
        intermediate lags where occasional long glides dominate, and decaying if a
        central-limit-type Gaussianisation sets in at long lags). It is a compact summary of
        the shape of $P$, not a substitute for it.
  \item[Velocity autocorrelation $C(\tau)$.]
        $C(\tau)=\langle\mathbf{v}(t_0)\cdot\mathbf{v}(t_0+\tau)\rangle$ (often normalised by
        $C(0)$) is the dot product of the velocity with its later self: it is large and
        positive when, a delay $\tau$ later, the motion still points the same way, so the
        scale over which $C(\tau)$ decays \emph{is} the memory time of the heading. It controls
        the MSD through the Green--Kubo relation
        $\mathrm{MSD}(t)=2\!\int_0^{t}\!(t-\tau)\,C(\tau)\,\mathrm{d}\tau$: a slowly decaying
        $C$ produces a faster-growing MSD. We do not presume the motion ends in ordinary
        diffusion; rather, the functional form of $C(\tau)$ is the diagnostic -- an integrable
        $C$ relaxes to normal diffusion, a non-integrable (e.g.\ power-law) tail sustains
        persistent super-diffusion with no crossover, and a sign change reveals the
        anti-persistence expected from circling.
  \item[Turning-angle distribution.] The distribution of the angle $\theta$ between two
        successive horizontal displacement vectors along the trajectory -- the change of
        heading from one step to the next. A distribution peaked at $\theta=0$ means strong
        forward persistence (a correlated random walk); a flat distribution means memoryless
        reorientation. This is the microscopic origin of the persistence seen in $C(\tau)$ and
        the MSD.
  \item[First-passage and first-exit times.] The first-passage time $T(L)$ is the first time
        the horizontal distance from take-off reaches a value $L$ (the first-exit time, the
        first time the trajectory leaves a disc of radius $L$). For a self-similar transport
        process the characteristic time scales with the length as
        $\langle T(L)\rangle\sim L^{1/H}$, with dynamic exponent $z=1/H$, so the slope of
        $\log\langle T\rangle$ against $\log L$ gives an estimate of $H$ obtained in the space
        domain, independent of the MSD; the full distribution of $T(L)$ additionally exposes
        heavy tails coming from long trapping events.
  \item[Speed and vertical-velocity distributions.] The magnitude of the horizontal velocity
        and the vertical velocity. Basic kinematic characterization; the vertical velocity also
        separates climbing from sinking and feeds the segmentation below.
\end{description}

\paragraph{\textcolor{red}{Trajectory geometry?????}}
\begin{description}
  \item[Tortuosity.] How much a path winds, quantified by the ratio between the length actually
        travelled and the straight-line distance between its endpoints over the same interval
        ($1$ for a straight line, larger for a more convoluted path). It is a geometric
        \emph{efficiency} descriptor and does not map one-to-one onto a transport exponent --
        very different dynamics can share a tortuosity -- so we use it descriptively and to
        separate phases (close to $1$ in glides, large in search and climb) rather than as an
        estimator of $H$.
  \item[Fractal dimension.] The box-counting dimension $d_f$ of the trail, i.e.\ the exponent
        with which the number of boxes of size $\epsilon$ needed to cover the path grows as
        $\epsilon\!\to\!0$. For a self-similar walk it is tied to the transport exponent by
        \begin{equation}
          d_f=\min\!\big(1/H,\,2\big),
        \end{equation}
        so a ballistic line ($H=1$) gives $d_f=1$, a Brownian path ($H=1/2$) gives $d_f=2$,
        and the reported sub-ballistic soaring ($H\approx0.88$) gives $d_f\approx1.14$. The
        fractal dimension is thus the geometric counterpart of $H$ and a third, independent
        route to it, after the MSD slope and the return/first-passage scalings; agreement
        among the three is a strong internal check, disagreement is evidence of multi-scaling.
  \item[Radius of gyration.] The root-mean-square distance of the trajectory points from their
        centroid, i.e.\ the characteristic spatial extent of a flight. As a single number it
        sets the natural length scale of a flight (useful to normalise other observables);
        measured as it grows with elapsed time, $R_g(t)\sim t^{H}$, it offers yet another
        reading of the same exponent.
\end{description}

\paragraph{One number per flight, or a function of scale?}
The geometric quantities above can be read in two complementary ways. As a \emph{single
number per flight} -- one tortuosity, one $d_f$, one $R_g$ -- whose distribution over the
ensemble characterises the population and feeds the stratified comparisons. Or as a
\emph{function of scale or elapsed time} -- $R_g(t)$ as the trajectory unfolds, tortuosity
measured on sub-segments of growing duration $t$, the box count as a function of box size --
which is precisely where the transport exponents live ($R_g(t)\sim t^H$; the box-count slope
is $d_f$). The scalar is descriptive; the scale dependence is what links geometry back to the
anomalous exponent, and both views will be used.

\section{Segmentation into flight phases}
Building on the previous work of Vilpellet et al.~\cite{vilpellet2026}, each flight is
segmented into the three phases of the soaring cycle:
\begin{description}
  \item[Transition (glide).] The inter-thermal cruise: a near-straight, energy-efficient
        glide toward the next source of lift (sinking, large directed horizontal
        displacement). This is the directed ``step''.
  \item[Search.] The exploratory phase entered when no thermal is found at a comfortable
        altitude: gentle turns and probing manoeuvres to sample the air mass and locate the
        next thermal while limiting altitude loss. It mediates the passage from one step to
        the next trap.
  \item[Climb (thermalling).] Circling within rising air to regain altitude: persistent
        turning with small net horizontal displacement. This is the localized ``trap''.
\end{description}
That work performs the segmentation with a hidden Markov model whose observations are a
small set of \emph{binary} trajectory-derived features --- the sign of the vertical speed,
the persistence of the curvature angle, and indicators of trajectory straightness.

A first task of this thesis is to scrutinize this segmentation rather than take it for
granted: to assess whether those binary variables are adequate and robust on the larger,
more heterogeneous dataset assembled here, and whether the feature set or the model should
be refined, extended, or replaced. The segmentation method is therefore itself an object of
study, and the specific variables are deliberately left open at this stage. This
three-state description refines the two-state caricature underlying continuous-time random
walks and L\'evy walks.

\section{Phase-resolved observables}
The segmentation unlocks a second group of observables, defined per phase occurrence and
then as ensemble distributions. These are the building blocks of an intermittent-transport
description and, unlike those of Section~\ref{sec:obs-global}, cannot be obtained without
first segmenting the flight.
\begin{description}
  \item[The waiting phase: climb and search.] Between two consecutive glides the flier
        advances little horizontally: it searches for a thermal and climbs in it. For a
        transport model these two phases together form the \emph{waiting time} between
        directed steps, so we treat them on an equal footing and build three distributions --
        the climb duration $\tau_c$, the search duration $\tau_s$, and their sum
        $\tau_w=\tau_s+\tau_c$, the total time spent essentially stationary between glides.
        The distinction is not cosmetic: Vilpellet et al.~\cite{vilpellet2026} report the
        search times to be broadly \emph{power-law} distributed while the climb times are
        closer to \emph{exponential}, so that the heavy tail of the combined waiting time
        $\psi(\tau_w)$ -- the quantity that actually decides whether trapping drives
        sub-diffusion -- would be inherited from the search phase. Whether this still holds on
        the larger dataset is itself one of the things to re-check once the segmentation is in
        place. We therefore report $\psi(\tau_w)$ together with the separate $\psi(\tau_c)$ and
        $\psi(\tau_s)$. Alongside the durations we record the climb kinematics (altitude gained
        $\Delta z$, mean climb rate, thermalling radius and period) and the search geometry
        (path length, tortuosity, net displacement) -- the search being the phase in which
        prior work locates the signature of pilot skill.
  \item[The directed step: glide (transition).] Each glide contributes a horizontal step of
        length $\ell$ and duration $\tau_g$, with ground speed, glide ratio (horizontal
        distance per unit altitude lost) and sink rate. Because a glide is close to ballistic,
        $\ell\approx v_g\,\tau_g$, step length and duration are expected to be nearly
        proportional; we test this directly through the joint law of $(\ell,\tau_g)$ -- the
        $\ell$--$\tau_g$ scatter, the conditional mean $\langle\ell\mid\tau_g\rangle$, and the
        spread of the ratio $\ell/\tau_g$ (the glide speed). Both marginals matter: the tail of
        the step-length distribution $\lambda(\ell)$ controls super-diffusive contributions, so
        we compute the distributions of $\ell$ and $\tau_g$ separately as well as their joint
        law.
  \item[Anatomy of a flight.] Number of thermals (climbs) per flight, inter-thermal distances,
        and the fraction of both time and along-track distance spent in each phase: the coarse
        composition that sets the weight of each phase in the global transport.
  \item[Angular diffusion between glides.] The turning angle between the heading of one
        transition segment and the next -- the segment-level counterpart of the turning angle
        of Section~\ref{sec:obs-global}. How fast successive glide directions decorrelate is
        the key ingredient of the minimal model below.
\end{description}

Two couplings then decide which transport model applies, and are stated here as explicit
objects of study. The first is the \emph{step--waiting coupling}: whether a long glide tends
to be followed by a long wait, i.e.\ whether $\ell$ and $\tau_w$ are correlated. A genuine
coupling is the defining feature of a L\'evy walk and is what separates it from a decoupled
CTRW, in which jumps and waiting times are drawn independently.

The second is the \emph{memory in the phase sequence}. A flight is a discrete sequence over
the three states $\{\text{transition},\text{search},\text{climb}\}$, and memory can hide in
two places. (i) The \emph{order} of the phases, summarised by the transition matrix
$P_{ij}=\Pr(\text{next}=j\mid\text{current}=i)$: a perfectly cyclic flight
($\text{T}\!\to\!\text{S}\!\to\!\text{C}\!\to\!\text{T}$) gives a near-deterministic matrix,
whereas aborted climbs, repeated searches or skipped phases appear as off-cycle probabilities
and reveal how stereotyped the soaring cycle actually is. (ii) The \emph{durations}, through
correlations between consecutive phase durations -- for instance whether a long climb tends to
precede a long glide. Both probe the \emph{renewal} assumption underlying the textbook CTRW
and L\'evy-walk results, in which each step and wait is independent of the history: if the
transition matrix is structured or the durations are correlated, the process is non-renewal
and those formulas must be amended. Measuring these is therefore what tells us whether a
memoryless model is admissible at all.

\section{Reproducing the reference results}
A first objective of the analysis, undertaken once the segmentation is in place, is to check
whether the enlarged---and eventually multi-source---dataset reproduces the quantitative and
graphical results of Vilpellet et al.~\cite{vilpellet2026}. Two figures are the main targets.

The first is the per-phase conditional MSD (their Fig.~3): the transition segments should
stay ballistic ($\alpha=2$, i.e.\ $H=1$) across time lags; the search phase should be
strongly sub-diffusive ($H\approx0.3$ for paragliders and hang gliders, $\approx0.2$ for
sailplanes); and the climb phase should be quasi-ballistic at long lags but with an effective
velocity about two orders of magnitude smaller, including the transient anti-correlation near
the $\sim\!30$\,s thermalling period.

The second is the set of per-phase descriptive statistics (their Fig.~4): the glide ratios
($\approx 8.5$, $10$, and $25$ for paragliders, hang gliders, and sailplanes), the
horizontal-velocity statistics, the heavy-tailed distribution of transition times, the
search-time fraction and the effective search radius, the vertical climb velocities, the
thermalling radii, and the thin, near-exponential distribution of climb times---together with
the global time budget ($\approx\!60\%$ transition, $\approx\!35\%$ climb, the remainder
searching) and the global sub-ballistic scaling ($H\approx0.88$).

Recovering these results would validate the acquisition pipeline and the segmentation and
would test the universality of the reported behaviour on a substantially larger sample;
systematic departures would be equally informative, pointing to dataset- or method-dependent
effects to investigate. This comparison is therefore the bridge between the present
data-acquisition stage and the original analysis that follows.

\section{Transport analysis}
Using these observables, test for anomalous transport: estimate the MSD scaling exponent
$\alpha$, fit the tails of $\psi(\tau)$ and $\lambda(\ell)$ with principled methods (e.g.\
maximum-likelihood power-law fits with goodness-of-fit tests), and perform model selection
between the continuous-time random walk and the L\'evy-walk
frameworks~\cite{metzler2000,zaburdaev2015} -- using the step/waiting-time coupling to tell
them apart. Because logging devices, sites, gliders and weather differ, results will be
stratified (by season, site, glider class, and data source) and trajectories normalized,
and finite-size, censoring and sampling biases will be controlled throughout.

\section{Modelling and simulation}
The aim is a \emph{minimal} stochastic model that reproduces the asymptotic anomalous
scaling reported for soaring (a Hurst exponent $H \approx 0.88$) with as few ingredients as
possible. A preliminary working hypothesis---to be made precise, and possibly revised, once
the empirical phase statistics are available---is that the transport is carried mainly by
the transition (glide) phases. The model would therefore resolve the transition segments
explicitly while merging search and climb into a single effective non-transition state, so
that the soaring cycle reduces to a sequence of directed glides interrupted by
reorientations. Anomalous transport would then emerge from the directed transition segments
combined with an \emph{angular diffusion} of the heading from one transition to the next:
successive glide directions decorrelate gradually rather than abruptly, producing a
correlated random walk whose persistence controls the long-time exponent. Simulated
ensembles would be compared with the data to test whether these ingredients suffice. These
are baseline ideas, to be refined---or changed---as the analysis progresses.

\section{Tooling}
Mirror the acquisition package with an analysis package, and let the figures and tables of
this document be generated automatically from the analysis outputs -- as the dataset
statistics already are -- so that this document stays in step with the work and remains a
reliable basis for the thesis and for presentations.
