\chapter{The dataset}
\label{impl:dataset}
\emph{Mirrors Chapter~\ref{ch:dataset} section by section, opening with the data
acquisition (Sec.~\ref{sec:acquisition}) and including the pre-processing and preliminary-
characterization sections (Secs.~\ref{sec:preproc},~\ref{sec:prelim}) -- the cleaning and
first-look steps are part of describing the dataset, not of the planned analysis in
Chapter~\ref{ch:next-steps}.}

\input{appendices/impl/C1-acquisition}

\section{The IGC tracklog format}
\label{impl:igcformat}
The parser (\path{soaring.analysis.igc.parse_igc}) checks, at decode time and
independently of any flight-dynamics judgement: that the time is a valid 24-hour
\texttt{hh:mm:ss}; that the arc-minutes field of latitude and longitude is
\texttt{<\,60} (the \path{DDMM.mmm} encoding is undefined otherwise); that the
hemisphere letter is one of the two valid characters; and that the decoded
latitude/longitude fall within $[-90,90]$/$[-180,180]$ degrees. A record failing any of
these is dropped as unparseable, on the same footing as a record with the wrong length or
a non-numeric field. This format-validity check is implemented now and covered by unit
tests; the fix-level dynamics-plausibility checks it is distinct from
(Sec.~\ref{impl:fixlevel}) are designed and thresholded but not yet implemented.

\section{Flight metadata and the catalog}
\label{impl:catalog}
Both \path{catalog.csv} and \path{seasons_index.csv} are regenerated from the
archived season XMLs on demand, via the acquisition CLI's \texttt{build-catalog}
subcommand (Sec.~\ref{impl:acqpipeline}) -- never hand-edited.

\section{Glider categories}
\label{impl:glider}
Label normalisation (not yet implemented) will merge \path{aile_class} variants and
recover a blank or placeholder (\enquote{0}) class from the wing model (\texttt{aile})
where possible, at the fix-level/pre-processing stage (Sec.~\ref{impl:fixlevel}).

\section{Coverage and statistics}
\label{impl:coverage}
\emph{To be populated: season-table provenance.}

\section{Trajectory pre-processing}
\label{impl:preproc}
\emph{Mirrors Sec.~\ref{sec:preproc} and its subsections below.} The Welch/PSD
computational procedure used to compare altitude channels (Sec.~\ref{impl:altchannel})
is the same one Sec.~\ref{sec:savgol} will reuse, applied to $E,N,U$ instead of altitude
alone; \ref{impl:savgol} should cross-reference \ref{impl:altchannel} for that shared
procedure rather than restate it, once populated.

\subsection{Choice of altitude channel}
\label{impl:altchannel}

\paragraph{PSD sample (Figure~\ref{fig:altnoise}a--c).} A seeded random subsample of
\num{3000} flights per discipline -- larger than a spectral \emph{shape} alone would need,
because the high-frequency barometric curve is visibly under-averaged (ragged) at a few
hundred flights and smooths out only with a few thousand. Panel~(a)/(b) use one
representative flight (the one with the most fixes in the subsample); its measured offset
(panel~b) is annotated on the panel itself, not restated here.

\paragraph{Welch/PSD computational procedure (panel c).} IGC loggers record one fix per
second nominally, but timestamps are not guaranteed to be perfectly regular (a missed
fix, a logger writing at \SI{1.05}{\second} instead of \SI{1.00}{\second}); Welch's
method (Appendix~\ref{app:psd}) requires a uniformly sampled signal, so the following
runs on every flight before the transform.
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Modal sampling period.} For each flight the median inter-fix interval is
        computed and rounded to the nearest integer second. The value that appears most
        often across all flights in the subsample (the \emph{mode}) is taken as the
        common target $\Delta t$ (in practice $\Delta t=\SI{1}{\second}$ for both
        disciplines, giving $f_s=\SI{1}{\hertz}$ and $f_{\text{Nyq}}=\SI{0.5}{\hertz}$).
  \item \textbf{Flight selection.} Only flights whose own median period lies within
        \SI{0.25}{\second} of the modal $\Delta t$ are retained, so pooled spectra share
        one frequency axis; flights with fewer than $N_{\text{seg}}=256$ fixes
        ($\sim$\SI{4}{\minute} at $\Delta t=\SI{1}{\second}$) are also excluded, since at
        least one full segment is needed. Of the \num{3000} sampled flights per
        discipline, this leaves \num{1641} paragliders but \num{727} hang gliders:
        paragliders are predominantly \SI{1}{\second} loggers
        (\StatScanParaDtOneSecPct\,\%, Sec.~\ref{impl:uniform}), so the population mode is
        a good match for most of them, whereas hang gliders split across several native
        rates ($1$, $2$, $5$, $\geq\!8\,\text{s}$, Figure~\ref{fig:sampling}) and only
        their \SI{1}{\second}-logger minority survives this cut.
  \item \textbf{Uniform resampling.} The altitude series $z(t)$ is linearly interpolated
        onto a grid $t_0,t_0+\Delta t,t_0+2\Delta t,\ldots$ spanning the flight, without
        introducing frequency content beyond $f_{\text{Nyq}}$.
  \item \textbf{Welch on the resampled series.} \path{scipy.signal.welch} is called
        with $f_s=1/\Delta t$, $N_{\text{seg}}=256$, 50\% overlap, a Hann window, and
        linear detrending per segment.
  \item \textbf{Robust ensemble estimator.} The per-flight PSDs are kept individually
        (per discipline and channel) and reduced across flights by the per-frequency
        \emph{median}, with a shaded \numrange{10}{90} percentile band; panel~(c) then
        pools the two disciplines, whose medians coincide. The median, not the arithmetic
        mean, because these per-flight spectra are heavy-tailed: a mean is dominated by a
        few flights and swings by one to two orders of magnitude between samples -- for the
        hang-glider barometric floor, from \num{0.8} to \num{93}\,\si{\metre\squared\per\hertz}
        between an $n=\num{114}$ and an $n=\num{727}$ draw -- whereas the median is stable.
        Because $\Delta t$ is the same for every retained flight by construction, the
        figure's Nyquist dotted line is a single fixed value.
\end{enumerate}

\paragraph{Reading the band (panel c).}
\begin{description}
  \item[Why the medians coincide.] For the typical flight both channels are dominated at
        high frequency by the metre resolution to which the IGC format rounds every
        altitude: the white-noise floor of a \SI{1}{\metre} quantiser,
        $\tfrac{1}{12}/f_{\text{Nyq}}\approx\SI{0.17}{\metre\squared\per\hertz}$, sits
        just below where the measured medians level off
        ($\approx\SI{0.3}{\metre\squared\per\hertz}$).
  \item[Where the channels part.] In the tail: at the 90th percentile the
        high-frequency GNSS floor is \num{4}$\times$ the barometric one for paragliders
        and \num{108}$\times$ for hang gliders (the worse GNSS multipath of that fleet,
        and the reason its band is wider), and \SIrange{13}{20}{\percent} of flights
        carry a GNSS floor above three times their own barometric one.
  \item[Consequence.] This heterogeneity, rather than a uniform gap, motivates the
        choice of the barometric channel (Sec.~\ref{sec:altchannel}).
\end{description}

\paragraph{Fallback-rate census (panel d).} A full census, not a sample: the same full
parse of every track is needed anyway for the flight-level filtering diagnostics of
Sec.~\ref{impl:flightfilter} (Figure~\ref{fig:preproc}), so the barometric-presence
fraction is read off that one scan rather than measured separately, at no extra cost. (A sized random sample -- the standard proportion formula
$n=z^2p(1-p)/e^2$ -- would be the right tool if a full census were not already being
paid for elsewhere; it is not needed here.)

\paragraph{Presence is whole-file (panel d).} The barometric channel is all-or-nothing at the
file level, checked on the same census: when it is absent it reads exactly zero for
\StatScanParaBaroZeroPct\,\% (paragliders) / \StatScanHangBaroZeroPct\,\% (hang gliders) of
those flights, and when present it covers every fix for \StatScanParaBaroFullPct\,\% /
\StatScanHangBaroFullPct\,\% of them (the rest miss only a
handful of individual fixes, dropped one by one, Sec.~\ref{sec:fixlevel}). A logger either has
a pressure sensor for the whole track or none at all, so the \texttt{alt\_source} decision is
per flight, never per fix.

\subsection{Fix-level cleaning}
\label{impl:fixlevel}

This section gives the working definitions the narrative (Sec.~\ref{sec:fixlevel}) states in
words: the three detectors, their parameters, the order they run in, and how the whole step
is validated. The three families of Table~\ref{tab:cleaning} map onto three detectors of
increasing context: an absolute bound (no context), a robust local test (a short window of
context), and a structural rule (a whole run).

\paragraph{Status.} The format-validity rejection is implemented in
\path{soaring.analysis.igc.parse_igc} and covered by unit tests. The dynamics and structural
cleaning below is designed, and its absolute bounds are thresholded on the data
(Table~\ref{tab:cleaning}, Figure~\ref{fig:fixlevel}); the routine that walks a trajectory
applying all three detectors, and the robust-test and frozen-lock parameters it needs, are
the next step. The parameter names introduced here are the config keys that step will add.

\paragraph{Robust local-outlier test (off-the-trend defects).} For a fix $P_k$ at time
$t_k$:
\begin{enumerate}[label=(\arabic*), itemsep=1pt, topsep=3pt]
  \item take the fixes lying within $\pm w$ seconds of $t_k$ (a \emph{time} window, so it
        adapts to the native cadence and to irregular sampling), excluding $P_k$ itself;
  \item fit a robust local estimate of the position there -- the component-wise median of
        the windowed fixes, or a degree-one Theil--Sen fit -- and evaluate it at $t_k$;
  \item the residual $r_k$ is the great-circle distance from $P_k$ to that estimate; scale
        it by the robust local spread $\sigma_k = 1.4826\,\operatorname{MAD}$ of the
        windowed residuals (the factor makes $\sigma_k$ a Gaussian-consistent standard
        deviation~\cite{rousseeuw1993});
  \item flag $P_k$ when $r_k > k\,\sigma_k$ \emph{and} $r_k > \varepsilon_{\min}$.
\end{enumerate}
This is a Hampel identifier~\cite{hampel1974}: a median centre, a MAD scale, a threshold in
robust standard deviations; working defaults in Table~\ref{tab:fixlevel-params}. The
window must hold at least a handful of fixes (order five); a sparser neighbourhood falls
back to the absolute bounds alone.

Three points earn the test its place over the bare speed bound:
\begin{itemize}[itemsep=2pt, topsep=3pt]
  \item \emph{The floor $\varepsilon_{\min}$ is not cosmetic.} On a very smooth stretch
        $\sigma_k\to 0$, so without it any sub-metre wiggle would exceed $k\sigma_k$ and
        the test would chase GPS noise where there is no outlier.
  \item \emph{It is local, and locality buys attribution.} A speed belongs to the segment
        \emph{between} two fixes and cannot say which endpoint is at fault, whereas the
        residual belongs to one fix: when the absolute bound fires, the identifier
        attributes the defect to a single fix, and a burst is removed as one block -- a short burst of
        consecutive spikes inflates the windowed MAD only mildly, so it is still flagged,
        and is removed as the block whose bracketing neighbours rejoin within scale.
  \item \emph{It does not clip real manoeuvres.} A genuine sharp turn is drawn by
        consecutive, mutually consistent fixes, each well predicted by its neighbours and
        so of small residual; only an isolated point off the local trend scores large. The
        one failure mode -- a turn so tight it is under-sampled at the native cadence -- is
        rare at these rates, and the injected-defect test below measures it directly.
\end{itemize}
The flag itself is context-sensitive -- a \SI{30}{\metre\per\second} step out of a
\SI{12}{\metre\per\second} glide is flagged while the same step passes unremarked where it
is ordinary -- but a flag alone does not delete (next paragraph): a
flagged-but-physically-possible fix is \emph{kept}, its flag recorded as a per-flight
anomaly count alongside the integrity-gate counters. The identical test runs on the
horizontal position and, independently, on the adopted altitude channel, so a vertical
spike invalidates the altitude alone and the horizontal position stays (the channel
decoupling of Sec.~\ref{sec:fixlevel}).

\paragraph{From flag to action.} The identifier only
\emph{flags}; it never repairs. A flagged horizontal fix is \emph{deleted} as a node -- not
overwritten with the local median -- so that all reconstruction is done by the single
resampling step (Sec.~\ref{impl:uniform}), under one policy and one accounting, rather than by
a second, in-place imputation that would then feed the resampler as if it were data. This is
the \emph{Hampel identifier}~\cite{hampel1974}, the detection half of the eponymous filter, not
the median-substituting filter itself: the local median/Theil--Sen enters only as the reference
for the residual $r_k$, never as an output value.

\begin{description}
  \item[Position.] Deletion requires two conditions: the fix must be flagged \emph{and}
        its implied in-and-out speed must exceed the absolute $v_{xy}$ bound. The gate
        narrows the identifier's deletions to what the bound already deems impossible;
        the one benign false positive, a genuine under-sampled tight turn (flagged, but
        flying at ordinary speed), is kept, and its bounded residual effect is absorbed
        by the smoother.
  \item[Discontinuity.] When no bounded block restores continuity -- a position
        discontinuity, such as a re-acquisition offset, where every removal still leaves
        an impossible step -- the scan stops deleting and marks a split at the step,
        handled like a long gap (Sec.~\ref{impl:uniform}): both sides are genuine, only
        the transition is unknown.
  \item[Altitude.] A flagged altitude is marked missing on the flag alone: a dropped
        altitude is a deferral, restored by the audited interpolation at resampling, so
        the stakes are far lower than for a position node. On a GNSS-fallback flight the
        absolute $|v_z|$ bound (calibrated on the barometric channel) does not apply,
        and this test is the only vertical cleaner.
\end{description}
This is the \emph{default to keep} rule of Sec.~\ref{sec:fixlevel}: removal requires
positive, corroborated evidence.

\paragraph{Frozen-lock detection (on-the-trend).} A run is a lost-lock episode only under the
rule of Eq.~\eqref{eq:frozenlock} -- spatially collapsed, \emph{and} witnessed (barometrically
flat or byte-identical; the declared flags only where no barometer exists), \emph{and} long
enough -- and the pass defaults to keeping the data otherwise.
\begin{itemize}[itemsep=2pt, topsep=3pt]
  \item \emph{Extent.} The run's bounding diameter -- the maximum great-circle distance
        between \emph{any} two of its fixes, not the fix-to-fix step -- stays below
        $\varepsilon$ (a few times the GPS noise, $\sim$\SIrange{10}{15}{\metre}); a
        circling climb spans its disc,
        $2R\approx\SIrange{40}{100}{\metre}\gg\varepsilon$, and is cleared on extent
        alone, whereas a step-based test would fire on every \SI{1}{\hertz} climb (step
        $v_{xy}\Delta t\approx\SI{10}{\metre}\sim\varepsilon$).
  \item \emph{Independent witness.} Over the run the barometric altitude must also be
        flat ($|\Delta z_{\mathrm{baro}}|$ below a few metres), since a real thermalling
        climb gains tens of metres in that time on a sensor independent of the GNSS, so a
        horizontally-still run whose baro altitude is still climbing is a tight climb and
        is kept. $\Delta z$ is taken robustly (medians of the run's ends), so an altitude
        spike inside the run -- not yet cleaned, since the altitude pass comes last --
        cannot fake a climb.
  \item \emph{Corroboration.} Of the declared signatures, only the
        \emph{byte-identical} repeat may override the witness -- a run of byte-identical
        coordinates under a climbing barometric altitude is a freeze that happened
        \emph{during} a climb (no real circling wing rewrites the same coordinates), and
        is cut, the positions being invented even where the height gain is real. The
        \texttt{V} flag and the zeroed GNSS altitude do \emph{not} override a live
        barometer: they certify a degraded GNSS \emph{altitude}, not a frozen horizontal
        fix (a \texttt{V} fix may still carry a usable position,
        Sec.~\ref{sec:igcformat}), so on a barometric flight they are recorded as
        corroborating diagnostics only, and become decisive only on a GNSS-fallback
        flight, where Eq.~\eqref{eq:frozenlock} substitutes them for the missing
        barometric witness.
  \item \emph{Duration.} The run must span at least $\tau_{\mathrm{freeze}}$, set to a few
        thermalling periods ($\gtrsim\SIrange{60}{90}{\second}$), so that a long, slow,
        tight climb is never mistaken for a freeze.
\end{itemize}
Only then is the run marked as a data gap; the split it induces is
realized at the uniform-sampling stage (Sec.~\ref{impl:uniform}), on the same footing as a
native long gap. On a GNSS-fallback flight the barometric witness is unavailable, so there the
extent and flag tests must both hold outright and, in any doubt, the run is kept -- the safe
direction, since keeping a rare true freeze costs far less than splitting a real climb.

\paragraph{Duplicate and backward timestamps (on-the-trend).} The parser already rebuilds a
monotone clock: it unwraps the UTC-midnight roll-over (a drop of $\sim$\SI{86400}{\second}
advances the day count) and clamps residual backward jitter with a running maximum
(Sec.~\ref{impl:igcformat}). The redesign changes one behaviour: a backward step that is
\emph{not} a roll-over is no longer silently clamped but removed -- the fix with the corrupt
time is deleted as a node, just as a fix with a corrupt position is. A corrupted timestamp
is recorded as a defect rather than silently repaired. Which fix carries the corrupt time
is decided by minimal removal: delete the smallest set whose removal leaves a strictly
increasing clock (the complement of the longest increasing subsequence), so a single fix
with a forward-jumped clock is itself removed, rather than condemning every later fix
against a running maximum. This requires moving the clamp out of the parser:
\path{parse_igc} keeps the roll-over unwrap (a format concern, needing the raw time of day)
but stops flattening backward jitter with the running maximum -- flattened, the evidence the
cleaning must act on no longer exists downstream.

Fixes sharing one UTC second are
collapsed to a single representative at that second's centroid position. This replaces the
provisional \enquote{keep the first} of the earlier design: keeping the first privileges an
arbitrary member and, for a genuine $\ge\!\SI{2}{\hertz}$ logger, would silently halve the
cadence, whereas the centroid is order-independent and the sub-second resolution it discards
is discarded anyway once the flight is resampled to a common $\Delta t$
(Sec.~\ref{impl:uniform}). One dilution case is accepted: if one member of a duplicate pair
is itself a position spike, the centroid halves its amplitude before the outlier pass (which
runs after the time defects) sees it; a large spike is still flagged at half size, and one
small enough to slip under $k\sigma_k$ after halving is within what the smoother absorbs.

\paragraph{Absolute bounds (out-of-range, the backstop).} The six numbers of
Table~\ref{tab:cleaning} are the \texttt{FixLevelThresholds} dataclass, loaded by
\path{soaring.analysis.preprocessing.load_preproc_config} from
\path{configs/preprocessing.yaml} (key \path{fix_level}); nothing is hard-coded in Python.
They are the context-free floor of the design, and are placed on the real per-fix
distributions (Figure~\ref{fig:fixlevel}) rather than guessed. The horizontal-speed bound is
set where the bin-to-bin ratio of a finely binned (\SI{2.5}{\metre\per\second}) histogram
settles at $\approx 1$, i.e.\ where the decay has actually stopped, not where it first visibly
slows ($\approx\SI{40}{\metre\per\second}$ for hang gliders, versus the adopted
\SI{55}{\metre\per\second}), since a shallow but real decreasing stretch can still be genuine
rare dynamics.

\paragraph{Diagnostic sample (Figure~\ref{fig:fixlevel}).} Computed by
\path{fix_level_distributions} and rendered by \path{make_fixlevel_diagnostics_figure}
(Table~\ref{tab:impl-repro}). Panels~(a)--(c) pool a seeded random sample of \num{15000}
paraglider flights (\num{126723499} consecutive-fix pairs, $\sim$\SI{73}{\second} to compute)
and the complete \num{\StatScanHangTracks}-flight hang-glider population (a full census
at this size,
\num{37915334} pairs, $\sim$\SI{25}{\second}) -- under two minutes total. The large sample is
deliberate: resolving the sparse artefact tail well enough to tell it from real dynamics
needs far more points than a headline statistic would. The $x$-range in (a)/(b) extends to
each discipline's own 99.99th percentile rather than cropping just past the bound; panel~(c)
keeps a narrower range, its cut sitting far out in an otherwise tight, unimodal distribution.
Panel~(b) bins one integer \si{\metre\per\second} per bin because barometric altitude is
logged to the metre: differencing at a \SI{1}{\second} cadence returns integer rates, and a
finer grid would alias them into a spurious comb. (Not every flight samples at
\SI{1}{\second}, Figure~\ref{fig:sampling}, and differencing over a longer native step
returns finer values (multiples of \SI{0.2}{\metre\per\second} at \SI{5}{\second}) that
partly fill the gaps once cadences are pooled: among \SI{1}{\second}-native fixes about
99\%\ land within \SI{0.05}{\metre\per\second} of an integer, against only 34--42\%\ at every
other cadence.)

\paragraph{Detector order and complexity.} The detectors run in a fixed order:
\begin{enumerate}[label=(\arabic*), itemsep=0pt, topsep=3pt]
  \item format validity, at decode time (Sec.~\ref{impl:igcformat});
  \item time defects: backward steps, then duplicates;
  \item position outliers (Hampel flag $+$ impossibility gate);
  \item frozen-lock runs;
  \item altitude: absolute bounds and altitude outliers.
\end{enumerate}
The order matters because a speed or a residual is only meaningful once the time base is
monotone and free of duplicates. The position-outlier and frozen-lock passes are one
left-to-right scan that carries the last accepted fix as an anchor: it never re-walks the
whole track, and it absorbs a run by extending the anchor across the flagged block --
tracking the block's running bounding box to test its extent against $\varepsilon$ -- so
the cost is $O(N)$ in the number of fixes.

\paragraph{Flight-level integrity gate.} The pass records how much it changed each flight (or
segment) -- \path{n_fix_raw}, \path{n_fix_clean} and \path{frac_interpolated} in
\path{flights_meta} -- counted over the airborne (trimmed) window only: junk in a ground
phase that trimming removes anyway must not condemn the flight. A flight whose
deleted-plus-reconstructed fraction exceeds a small $f$
(order \numrange{0.05}{0.10}) is dropped whole: beyond that point the trajectory is more
interpolation than measurement. This is the cleaning counterpart of the sampling-regularity
cut (Sec.~\ref{impl:uniform}).

\paragraph{Configuration.} The absolute bounds are already the \texttt{FixLevelThresholds}
group above. The remaining parameters are collected in Table~\ref{tab:fixlevel-params}: they
join that group in \path{configs/preprocessing.yaml} when the routine is built and, as with
every threshold in this pipeline, they live in the config, never in Python.

\begin{table}[htbp]
\centering
\small
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}llp{6.9cm}@{}}
\toprule
\textbf{Parameter} & \textbf{Working default} & \textbf{Role, and the scale that sets it} \\
\midrule
\multicolumn{3}{@{}l}{\itshape Robust local-outlier test (Hampel)}\\
\addlinespace[2pt]
$w$ & \SI{20}{\second} & half-width of the local time window; a few native
  intervals \\
$k$ & $5$ & flag threshold, in robust standard deviations $\sigma_k$ \\
$\varepsilon_{\min}$ & $\approx\SI{15}{\metre}$ & absolute residual floor; a few times the
  horizontal GPS noise, so a smooth stretch ($\sigma_k\to0$) is not chased \\
\addlinespace[6pt]
\multicolumn{3}{@{}l}{\itshape Frozen-lock rule (Eq.~\ref{eq:frozenlock})}\\
\addlinespace[2pt]
$\varepsilon$ & \SI{15}{\metre} & run-extent tolerance; a few times the GPS noise,
  $\ll 2R\approx\SIrange{40}{100}{\metre}$ of a circling climb \\
$\delta_z$ & $\approx\SI{5}{\metre}$ & barometric-flatness witness; a few times the
  sensor noise and quantisation, $\ll$ the $\gtrsim\SI{60}{\metre}$ a real climb gains
  over $\tau_{\mathrm{freeze}}$ \\
$\tau_{\mathrm{freeze}}$ & \SI{60}{\second} & minimum span of a cut run; two
  thermalling periods, so a slow, tight climb is never mistaken for a freeze \\
\addlinespace[6pt]
\multicolumn{3}{@{}l}{\itshape Flight-level integrity gate}\\
\addlinespace[2pt]
$f$ & $0.10$ & maximum deleted-plus-reconstructed fraction; past it the
  flight is being rebuilt, not cleaned, and is dropped whole \\
\bottomrule
\end{tabular}
\caption{Fix-level cleaning: parameters of the robust and structural detectors, with their
working defaults -- to be finalized by the injected-defect calibration, check~(1) of the
validation list below -- and the physical scale that sets each. They live as config keys
under \texttt{fix\_level} in \texttt{configs/preprocessing.yaml}; the absolute bounds of Table~\ref{tab:cleaning} already live there.}%
\label{tab:fixlevel-params}
\end{table}

\paragraph{Validating the cleaning.} The removed-fraction count of Figure~\ref{fig:fixlevel}
bounds only the false positives (over-cleaning). Five checks, part of the same planned step,
bound the false negatives and the bias:
\begin{enumerate}[label=(\arabic*), itemsep=2pt, topsep=3pt]
  \item \emph{Injected defects.} On tracks the cleaning leaves untouched, plant synthetic
        defects of known type and size -- position spikes of amplitude $a$, frozen runs of
        length $\tau$, backward-time glitches -- and measure the recovered fraction against
        size. This gives the detection floor (the smallest spike caught) and calibrates
        $k$, $w$, $\varepsilon$ and $\tau_{\mathrm{freeze}}$ to a target rate: the only
        direct measurement of what the cleaning misses.
  \item \emph{Downstream invariance.} Recompute the transport estimators -- the
        displacement-scaling exponent and the waiting-time-tail exponent -- over a grid of
        cleaning strengths, and take the operating point on the plateau where they stop
        moving; movement under stricter cleaning is the signature of a cut biting into
        signal.
  \item \emph{Detector nesting.} The fixes flagged by the absolute bound should nest
        inside those flagged by the local test (the impossible is a subset of the locally
        anomalous); a large disagreement points to a defect class neither was designed for
        -- a cheap guard.
  \item \emph{Per-phase removal rates.} The error this pipeline most wants to avoid is
        losing a slow-phase fix. Once the segmentation is in place
        (Ch.~\ref{ch:next-steps}), the removal and split rates are measured \emph{per
        phase} and required to be flat across transition, search and climb rather than
        enriched in the two slow phases; and synthetic tight climbs -- circles at flying
        speed, over a range of radii and wind drift -- planted in otherwise clean tracks
        confirm directly that the frozen-lock detector never splits a real climb. This is
        the false-\emph{positive} complement of check~(1), and its natural readout is the
        tail of $\psi(\tau_w)$: a cut biting into climb or search would thin that tail
        before it moved anything else.
  \item \emph{Isolation of the vertical cuts.} The fixes cut by
        $|v_z|>\SI{13}{\metre\per\second}$ must be \emph{isolated}, not consecutive: a
        sustained spiral dive holds \SIrange{15}{20}{\metre\per\second} of real sink, so a
        run-shaped excess is the signature of a genuine manoeuvre -- calling for a raised
        bound or a coherent-run exemption rather than censoring -- whereas an isolated
        super-threshold difference, bracketed by ordinary rates, is a sensor error and
        safe to cut.
\end{enumerate}

\paragraph{Reproduce.} Table~\ref{tab:impl-repro}, row \path{generate_preproc_figure.py} (one
run also regenerates Figures~\ref{fig:preproc}, \ref{fig:gaps} and~\ref{fig:sampling} from a
shared full-census scan; the fix-level panels use a separate sampled pass).

\subsection{Trimming of ground phases}
\label{impl:trimming}
The take-off speed (\SI{5}{\metre\per\second}) and sustained duration (\SI{30}{\second})
are the \texttt{TrimmingThresholds} dataclass, loaded from
\path{configs/preprocessing.yaml} (key \texttt{trimming}); as with every threshold in
this pipeline, nothing is hard-coded in Python.

The interior-ground guard of Sec.~\ref{sec:trimming} keeps
\texttt{interior\_ground\_s} ($T_{\mathrm{ground}}$, \SI{600}{\second}) and
\texttt{ground\_flatness\_m} (\SI{5}{\metre}) in the same group. A detected stint is excised and handed to the split
machinery of Sec.~\ref{impl:uniform}, so it needs no removal semantics of its own. Below
$T_{\mathrm{ground}}$ the same predicate is stored as a list of suspect intervals per
flight; the segmentation later maps them onto the phases they touch, and the
$\psi(\tau)$ sensitivity re-fits consume the flag.

\subsection{Flight-level filtering}
\label{impl:flightfilter}
The duration window ($\SI{40}{\minute}\leq T\leq\SI{16}{\hour}$), the path floor
($L\geq\SI{10}{\kilo\metre}$) and the altitude-activity floor (\SI{75}{\metre} on the
adopted channel) are the \texttt{FlightLevelThresholds} dataclass, from
\path{configs/preprocessing.yaml} (key \path{flight_level}); the census measurements
behind the 30-to-10 km change and the upper duration bound are in the narrative
(Sec.~\ref{sec:flightfilter}).

\paragraph{Diagnostic sample (Figure~\ref{fig:preproc}).}
\begin{description}
  \item[Scope.] The full census, every parsed track, no sub-sampling:
        \num{\StatScanParaTracks} paraglider and \num{\StatScanHangTracks} hang-glider
        tracks, from the \path{load_or_scan_tracks} scan; the rendering script is in
        Table~\ref{tab:impl-repro}.
  \item[Provenance.] The totals, like every census number in this document, are
        generated macros quoting the cached scan (\path{generate_census_stats.py}); the
        live archive grows by a trickle, so they can trail the download counts of the
        season tables until the cache is refreshed.
  \item[Headline medians.] Duration and path length:
        \SI{\StatScanParaMedianDurH}{\hour}/\SI{\StatScanParaMedianPathKm}{\kilo\metre}
        for paragliders,
        \SI{\StatScanHangMedianDurH}{\hour}/\SI{\StatScanHangMedianPathKm}{\kilo\metre}
        for hang gliders; the latter fly faster and farther.
  \item[One pass.] The scan is cached at \path{<data_root>/derived/track_scan.parquet}
        and also feeds Figures~\ref{fig:gaps} and~\ref{fig:sampling}.
\end{description}

\paragraph{Reproduce.} Table~\ref{tab:impl-repro}, row \path{generate_preproc_figure.py}.

\subsection{From geographic to local Cartesian coordinates}
\label{impl:enu}
No implementation content: this step is pure coordinate geometry with no configurable
parameters, and both of its figures (Figure~\ref{fig:ellipsoid}, Figure~\ref{fig:enu-rotation})
are hand-drawn \textsc{TikZ} diagrams, not data plots.

\subsection{Uniform sampling within a flight}
\label{impl:uniform}
The sampling thresholds (key \texttt{sampling}, the \texttt{SamplingThresholds}
dataclass) act as follows.
\begin{description}
  \item[Split bound.] $g_{\max}=\min\big(\texttt{max\_gap\_factor}\times\Delta t,\,
        \max(\texttt{max\_gap\_seconds},\,2\,\Delta t)\big)$, pinned by the constraint
        chain of Sec.~\ref{sec:uniform}. Its audit is a per-flight count -- the census
        scan records \path{max_gap_ratio} and \path{dt_s}, so the absolute gap is their
        product -- and joins Figure~\ref{fig:gaps} when regenerated.
  \item[Missing fraction.] Acts per \emph{segment}: a segment more than a tenth
        uncovered is dropped alone, the flight only if no segment survives. The
        per-flight value in the census scan is the pre-split diagnostic, not the
        operational gate.
  \item[Segment gate.] \texttt{min\_segment\_duration\_s} (\SI{90}{\second}). It only
        needs to support the smoothing window and the within-segment statistics: a few
        smoothing timescales, never less than $w\,\Delta t$.
  \item[Measured vs interpolated.] A grid point counts as measured when a fix lies
        within half a native step of it, interpolated otherwise; the same tolerance
        decides whether a flight is uniform as recorded (no interpolated points) or was
        resampled (\path{was_resampled}).
\end{description} The bound is on the \emph{time base} and is the same for
every channel: an order-of-magnitude smaller $v_z$ buys no wider gap for the altitude, since
the interpolation error scales with curvature, not speed, and a wide altitude gap would erase
the climb--glide transitions that $v_z$ is there to resolve (Sec.~\ref{sec:uniform}).

\paragraph{Segment handling.}
\begin{description}
  \item[Origin and clock.] A segment keeps its parent flight's origin and clock. The ENU
        conversion precedes this step, so every fix already carries parent-referred
        coordinates and time; a split only assigns a \path{segment_id}.
  \item[Gates.] The flight-level cuts are not re-applied to a segment
        (Sec.~\ref{sec:uniform}); only the minimal-duration gate above applies, and a
        dropped segment does not drop its siblings.
  \item[Tables.] One row per fix in \path{fixes}, keyed
        (\path{flight_id}, \path{segment_id}); per-segment records (start/end, fix
        count, \path{frac_interpolated}, censoring flags) in a \path{segments} table
        beside \path{flights_meta}. No separate files; the ensemble MSD at lag $t$ reads
        $\mathbf{r}(t)$ wherever some segment covers $t$.
  \item[Censoring.] Any phase duration truncated at a segment boundary is flagged
        censored, for the $\psi(\tau)$ fits to exclude.
\end{description}

\paragraph{Per-channel fill.} Across a short gap the horizontal position is interpolated
shape-preservingly (monotone piecewise-cubic, e.g.\ \path{scipy.interpolate.PchipInterpolator})
and the barometric altitude linearly or by cubic spline (its within-phase smoothness makes the
choice immaterial). The result meets the completeness invariant of Sec.~\ref{sec:fixlevel}:
within each retained segment every grid time carries a defined $(x,y,z)$ -- deleted position
nodes and deferred (missing) altitudes both restored here -- so the segmentation is never handed
a hole; long gaps are split, not filled. Each filled grid point is flagged in an
\path{interpolated} boolean column of \path{fixes}, and \path{frac_interpolated} records
the per-segment fraction for the integrity gate.

\paragraph{Figures~\ref{fig:gaps} and~\ref{fig:sampling}.} Both rendered by
\path{make_gap_diagnostics_figure}/\path{make_sampling_figure} from the same
full-census scan as Figure~\ref{fig:preproc} (Table~\ref{tab:impl-repro}, row
\path{generate_preproc_figure.py}) -- no separate parse needed.

\paragraph{Gap-ratio discreteness (Figure~\ref{fig:gaps}).} Checked directly: only
\StatScanParaGapIntPct\,\%/\StatScanHangGapIntPct\,\% (para/hang) of largest-gap-ratio
values are exact integers, the rest a
genuinely continuous scatter; the integer peaks visible for hang gliders sit around
$k=3,6,10,15$.

\paragraph{Native-rate discreteness (Figure~\ref{fig:sampling}).} Checked directly:
$\Delta t$ lands on an exact whole second for \StatScanParaDtIntPct\,\% of paragliders
and \StatScanHangDtIntPct\,\% of hang
gliders. Per-rate breakdown: paragliders mostly (\StatScanParaDtOneSecPct\,\%) at
\SI{1}{\second}; hang
gliders spread across \SI{1}{\second} (\StatScanHangDtOneSecPct\,\%), \SI{5}{\second}
(\StatScanHangDtFiveSecPct\,\%), \SI{10}{\second}
(\StatScanHangDtTenSecPct\,\%), \SI{2}{\second} (\StatScanHangDtTwoSecPct\,\%), and a
long thin tail beyond \SI{11}{\second}.

\subsection{Smoothing and differentiation}
\label{impl:savgol}
Not yet implemented. Three decisions are pinned for the implementation: the filter runs
per segment (never across a boundary) with \path{scipy.signal.savgol_filter} in
\texttt{mode='interp'}, so the terminal half-windows are fitted on the edge window rather
than padded; edge samples inherit the segment-boundary caution of
Sec.~\ref{impl:uniform}; and the window is floored at $w=5$, the smallest a cubic fit
accepts, which for the few flights with $\Delta t\gtrsim\tau_c$ smooths slightly more
than the noise scale (those flights are stratified out of the $\Delta t$-sensitive
observables anyway). The polynomial order ($p=3$) and the three smoothing timescales --
\path{tau_c_horizontal_s} for $(E,N)$, \path{tau_c_vertical_baro_s} and
\path{tau_c_vertical_gnss_s} for $U$ -- are the \texttt{SavgolParams} dataclass in
\path{configs/preprocessing.yaml} (key \texttt{savgol}).

\paragraph{The measured timescales.} The knees were measured with
\path{estimate_savgol_timescales.py}, which applies the Welch procedure of
Sec.~\ref{impl:altchannel} to $E$, $N$ (local equirectangular metres) and to the altitude
channel each flight actually uses, on a seeded sample (908 horizontal flight-components,
719 barometric and 189 GNSS-fallback flights at \SI{1}{\hertz}). The knee rule: the floor
is the median PSD over \SIrange{0.35}{0.5}{\hertz}; $f_c$ is the lowest frequency whose
median PSD falls below three times it. All three knees land at
$f_c\approx\SI{0.2}{\hertz}$, so $\tau_c=\SI{5}{\second}$ for every channel: each
channel's typical floor is the metre and arc-minute quantization of the IGC format
(measured $\approx0.60$ and $\approx\SI{0.26}{\metre\squared\per\hertz}$ against the
theoretical $q^2/12/f_{\mathrm{Nyq}}$ of $0.57$ and $0.17$), not receiver noise. The
expected asymmetry between the barometric and GNSS verticals shows in the spectrum's
\emph{level}, not in its knee, and even the 90th-percentile spectra move the knee by less
than a second. The two vertical keys stay, conditioned on \path{alt_source}, for the
noisy-minority refinement; the validation of Sec.~\ref{sec:savgol} (flat residual
spectrum, stability under one window step) is what confirms the values. The filter runs
on the resampled grid, so the small interpolated fraction (\path{frac_interpolated}) is
already in place when it reads its window.

\section{Preliminary characterization}
\label{impl:prelim}
Not yet implemented; see Sec.~\ref{sec:prelim}.
