\chapter{Power spectral density and Welch's method}
\label{app:psd}

This appendix explains the tool used in Chapter~\ref{ch:next-steps} to compare the two
altitude channels of an IGC file, namely the \emph{power spectral density} (PSD) estimated
with \emph{Welch's method}. It is written so the reader need not already know what these are.

\section*{What the power spectral density is}

Take a signal sampled at a fixed rate, here an altitude series $z_n=z(n\,\Delta t)$,
$n=0,\dots,N-1$, with sampling interval $\Delta t$ (so the sampling frequency is
$f_s=1/\Delta t$). Any such finite signal can be written as a sum of sinusoids of different
frequencies (its discrete Fourier transform). The \textbf{power spectral density} answers a
simple question: \emph{how is the ``power'' (the variance) of the signal distributed across
frequencies?} A slow drift puts its power at low frequency; rapid jitter puts its power at
high frequency. Formally the PSD $S(f)$ is the Fourier transform of the autocovariance of
the signal, and it is normalised so that the total area under it equals the variance of the
signal,
\begin{equation}
  \int_0^{f_s/2} S(f)\,\mathrm{d}f \;=\; \operatorname{Var}(z),
\end{equation}
with units of (metres)$^2$ per hertz. The upper limit $f_s/2$ is the \emph{Nyquist
frequency}: a series sampled every $\Delta t$ cannot resolve anything faster than one
oscillation per two samples. Reading a PSD is therefore reading \emph{where in frequency}
the signal lives: this is exactly what separates a slowly drifting but locally clean channel
(power concentrated at low $f$) from a channel with a high-frequency noise floor (power that
stays large up to the Nyquist frequency).

\section*{The periodogram and why it is noisy}

The naive estimator of $S(f)$ is the \textbf{periodogram}: take the discrete Fourier
transform $\hat z(f)=\sum_n z_n\,e^{-2\pi i f n\Delta t}$ and form $|\hat z(f)|^2$
(suitably normalised). It is unbiased but \emph{not consistent}: its variance does not
shrink as the record gets longer -- a longer signal gives more frequency points, each still
as noisy as before. Plotted directly it is a spiky, unreadable curve, from which one cannot
reliably compare two channels.

\section*{Welch's method: segment, window, average}

Welch's method turns the noisy periodogram into a usable estimate by \emph{averaging}. The
recipe is:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Segment.} Split the record into several shorter, overlapping segments (here
        of length $N_{\text{seg}}=256$ samples, the standard $50\%$ overlap).
  \item \textbf{Window.} Multiply each segment by a smooth taper (a Hann window) that falls
        to zero at its ends. This suppresses the spurious high-frequency power (\emph{spectral
        leakage}) that an abrupt segment boundary would otherwise inject.
  \item \textbf{Average.} Compute the periodogram of each windowed segment and average them.
\end{enumerate}
Averaging $K$ segments cuts the variance of the estimate by roughly a factor $K$, at the
cost of coarser frequency resolution (set by the segment length rather than the whole
record). The result is a smooth PSD estimate whose \emph{level} can be compared meaningfully
between two signals. This is the standard, robust spectral estimator; here it is provided by
\texttt{scipy.signal.welch}. Before the transform each segment is \emph{linearly detrended},
which removes the mean and the overall climb/drift ramp so that the estimate reflects the
fluctuations about the trend -- the part where sensor noise lives -- rather than the trend
itself.

\section*{Computational details for the IGC dataset}

IGC loggers record one fix per second nominally, but the timestamps are not guaranteed to
be perfectly regular: small gaps (a missed fix, a logger that writes at 1.05\,s instead of
1.00\,s) are common. Welch's method requires a \emph{uniformly sampled} signal. The
following procedure is applied to every flight before the transform.

\begin{enumerate}[label=(\roman*)]
  \item \textbf{Modal sampling period.} For each flight the median inter-fix interval is
        computed and rounded to the nearest integer second. The value that appears most often
        across all flights in the subsample -- the \emph{mode} -- is taken as the common
        target $\Delta t$ (in practice $\Delta t = 1\,\text{s}$ for both disciplines, giving
        $f_s = 1\,\text{Hz}$ and $f_{\text{Nyq}} = 0.5\,\text{Hz}$).
  \item \textbf{Flight selection.} Only flights whose own median period lies within
        $0.25\,\text{s}$ of the modal $\Delta t$ are retained, so that the pooled spectra
        all share one frequency axis. Flights with fewer than $N_{\text{seg}} = 256$ fixes
        (roughly 4 minutes at $\Delta t=1\,\text{s}$) are also excluded, since at least one
        full segment is needed. Of the \num{400} sampled flights per discipline (with a
        barometric channel and enough fixes), this leaves \num{212} paragliders but only
        \num{114} hang gliders: paragliders are overwhelmingly \SI{1}{\second} loggers, so
        the population mode is a good match for most of them, whereas hang gliders split
        across several native rates ($1$, $2$, $5$, $\geq\!8\,\text{s}$, Figure~\ref{fig:sampling}) and only their
        \SI{1}{\second}-logger minority survives this cut. The hang-glider PSD below is
        therefore an average over that minority, not the full hang-glider population --
        an honest reading of the panel, not a claim about every hang-glider logger.
  \item \textbf{Uniform resampling.} The altitude series $z(t)$ is linearly interpolated
        onto a grid $t_0, t_0+\Delta t, t_0+2\,\Delta t, \ldots$ spanning the flight. This
        replaces the mildly irregular original timestamps with a perfectly uniform series
        at the modal cadence, without introducing frequency content beyond $f_{\text{Nyq}}$.
  \item \textbf{Welch on the resampled series.} \texttt{scipy.signal.welch} is called with
        $f_s = 1/\Delta t$, $N_{\text{seg}} = 256$, 50\,\% overlap, a Hann window, and
        linear detrending per segment (as described above). The resulting
        $f$-axis runs from $0$ to $f_{\text{Nyq}} = f_s/2$, the last
        resolvable frequency.
  \item \textbf{Ensemble average.} The PSD of each qualifying flight is added to a running
        sum, separately for each discipline and channel (barometric / GNSS). The mean over
        all contributing flights is the curve shown in Figure~\ref{fig:altnoise}(c).
\end{enumerate}

Because the frequency axis is determined entirely by $\Delta t$ -- which is the same for
every retained flight by construction -- the dotted line in the figure marking
$f_{\text{Nyq}} = 1/(2\,\Delta t)$ is a single fixed value, not a flight-by-flight average.

\section*{Reading the altitude spectra}

In Figure~\ref{fig:altnoise}(c) the PSD is computed this way for the barometric and the GNSS
altitude of each flight and averaged over the sampled ensemble, at the common sampling
cadence. Two features carry the argument of Chapter~\ref{ch:next-steps}. First, both channels
agree at \emph{low} frequency, where the real vertical motion (climbs and glides) lives.
Second, towards the \emph{Nyquist} frequency the GNSS curve flattens into a high plateau --
a white-ish noise floor -- lying about two to three orders of magnitude above the barometric
curve, which keeps falling. That gap is the high-frequency GNSS jitter, precisely the band
that a time derivative (vertical velocity) or a displacement increment amplifies, and it is
the quantitative reason the vertical dynamics are taken from the barometric channel.
