% =============================================================================
%  Logbook -- Anomalous transport in soaring flights
%
%  A running record of the succession of events (what was done, what worked, what
%  did not, and WHY), as a complement to the state document in ../thesis.
%
%  The "Timeline" section is auto-generated from `git log`; the "Narrative notes"
%  are written by hand and hold the reasoning the timeline cannot.
%
%  Build:  make logbook   (or: latexmk -pdf logbook.tex)
% =============================================================================
\documentclass[11pt,a4paper]{article}

\ifdefined\pdfinfoomitdate\pdfinfoomitdate=1\fi
\ifdefined\pdftrailerid\pdftrailerid{}\fi

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[english]{babel}
\usepackage[a4paper,margin=2.5cm]{geometry}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

% Compact, scannable sub-lists inside a dated entry: bold run-in label, no wasted
% vertical space. Used throughout "Narrative notes" instead of dense wall-of-text
% paragraphs.
\setlist[description]{leftmargin=1.6em, itemsep=3pt, topsep=3pt, parsep=0pt}

\title{Logbook --- Anomalous transport in soaring flights}
\author{Matteo Di Sante}
\date{Working log}

\begin{document}
\maketitle

\noindent\emph{Working notebook. The timeline is regenerated from the git history;
the narrative notes are written by hand.}

\section{Timeline (auto-generated from git)}
\input{generated/timeline}

\section{Narrative notes}

\paragraph{2026-06 --- Paraglider data acquisition complete.}
The full FFVL paraglider CFD archive (\texttt{parapente.ffvl.fr}, seasons 1999--2000 to
2025--2026) has been scraped: about 203{,}000 flight records, of which $\sim$186{,}000
carry a GPS track. All 186{,}052 available tracks were downloaded.
\begin{description}
  \item[What worked] Impersonating a Chrome TLS fingerprint (\texttt{curl\_cffi}) reliably
        passes the FFVL Cloudflare challenge; the resumable, atomic, validated download
        survived many interruptions without corruption.
  \item[Gotchas] 173 tracks could not be obtained --- 172 because the server returns
        something that is not a valid IGC (broken or placeholder links on FFVL's side, so
        a retry does not help), and 1 due to a transient Cloudflare challenge (worth a
        later retry). On macOS/exFAT the OS scatters \texttt{.\_*} sidecar files next to
        every written file; a \texttt{clean} step removes them.
\end{description}

\paragraph{2026-06 --- Integrity verified.}
A full on-disk re-scan (\texttt{verify}) confirmed that all 186{,}052 downloaded
files are structurally valid IGC: no truncated files, no leftover \texttt{.part}, no read
errors. The verification logic was promoted from a throwaway script to a first-class
\texttt{soaring-para verify} subcommand (reusing the download-time validator), with tests.

\paragraph{2026-06 --- Robustness of the data path.}
The external disk remounted under a different name (\texttt{HDD\_DISANTE} $\to$
\texttt{SSD\_DISANTE}), which would have silently broken a hard-coded path. Fix: removed
every machine-specific path from the repo (placeholder + \texttt{SOARING\_PARA\_DATA\_ROOT}
environment variable) and added a start-up guard that aborts with a clear, actionable
message when \texttt{data\_root} is unset/unmounted --- and verified that \texttt{/Volumes}
is not user-writable, so a wrong path can never cause a silent download to the wrong place.

\paragraph{2026-07 --- Hang-glider data acquired; naming corrected.}
\begin{description}
  \item[Second source added] The hang-glider CFD (\texttt{delta.ffvl.fr}, seasons
        2001--2002 to 2025--2026), sharing the acquisition code
        (\texttt{soaring.acquisition.ffvl}) via a second CLI entry point
        \texttt{soaring-delta} (config \texttt{delta\_download.yaml}, env var
        \texttt{SOARING\_DELTA\_DATA\_ROOT}). Data: $\sim$9{,}300 flights, 6{,}716 tracks
        downloaded, 30 failures (broken server-side links, same pattern as paragliders).
        Stored at \texttt{/Volumes/SSD\_DISANTE/hang\_gliders/delta\_cfd\_igc/}.
  \item[Naming corrected] The paraglider CLI was renamed from the generic
        \texttt{soaring-ffvl} to \texttt{soaring-para} (and
        \texttt{SOARING\_FFVL\_DATA\_ROOT} $\to$ \texttt{SOARING\_PARA\_DATA\_ROOT}), and
        its config from \texttt{ffvl\_download.yaml} to \texttt{para\_download.yaml}, to
        avoid ambiguity now that both sources are present.
\end{description}

\paragraph{2026-07 --- Hang-glider catalog built; datasets reorganized symmetrically.}
Ran the full pipeline on the hang-glider data (\texttt{soaring-delta clean}, \texttt{verify},
\texttt{build-catalog}, \texttt{status}), which so far had only been downloaded: 9{,}259
flights, 6{,}746 with a track, 6{,}716 downloaded and all verified (0 integrity failures),
30 unrecoverable server-side links.
\begin{description}
  \item[Data layout] The local \texttt{data/} folder now mirrors the SSD --- one directory
        per discipline (\texttt{paragliders/}, \texttt{hang\_gliders/}), each with a
        versioned \texttt{seasons\_index.csv} and a git-ignored \texttt{catalog.csv}.
  \item[Thesis symmetry] \texttt{generate\_stats.py} now reads both indices and emits
        per-discipline macros (\texttt{\textbackslash StatPara*},
        \texttt{\textbackslash StatHang*}) and two per-season tables; the first three
        thesis chapters were made symmetric across both disciplines.
\end{description}

\paragraph{2026-07 --- Altitude channel decision: barometric.}
Decided to take the vertical dynamics from the \emph{barometric} altitude rather than the
GNSS altitude, diverging from the reference pipeline (J\'er\'emie's code derives $z$ and
$v_z$ from \texttt{AltGNSS}, reading but not using \texttt{AltPre}).
\begin{description}
  \item[Rationale] The pressure altitude has sub-metre relative precision and low
        high-frequency noise, its errors being slow (offset, drift) and killed by
        differencing, whereas the GNSS vertical carries a high-frequency noise floor that
        derivatives amplify. Confirmed empirically with a new noise diagnostic --- an IGC
        \texttt{B}-record parser (\texttt{soaring.analysis.igc}) and an altitude-noise
        analysis (\texttt{soaring.analysis.altitude\_noise}) whose Welch power spectra
        show the GNSS floor two to three orders of magnitude above the barometric one
        near Nyquist (thesis appendix on the method).
  \item[Consequences] The \texttt{A}/\texttt{V} validity flag is subsumed by the
        missing-altitude check on the adopted channel (a \texttt{V} fix keeps position +
        baro on a baro flight, but is dropped on a GNSS-fallback flight); one altitude
        channel per trajectory, no baro/GNSS splicing; flights with no pressure sensor
        fall back to unfiltered GNSS, tagged by a per-flight \texttt{alt\_source} column;
        the trimming threshold is on horizontal speed only ($v_{xy}>5$\,m/s, sustained);
        cleaning/trimming run on the raw geographic coordinates (great-circle speeds),
        with the ENU conversion done \emph{last}, origin at take-off.
\end{description}

\paragraph{2026-07 --- Baro-absence fraction: sized sampling, not a full sweep.}
A full-population parse of the paraglider archive (186{,}052 files) for the
baro-channel-absence figure took an unacceptable $\sim$45\,min serially; parallelising
across 8 processes cut this to $\sim$15\,min, still too slow to run routinely.
\begin{description}
  \item[Fix] A properly justified \emph{simple random sample}: the standard proportion
        sample-size formula ($n=z^2p(1-p)/e^2$, conservative $p=0.5$) gives $n=2401$ for a
        target 95\%-confidence margin of error of $\pm2$ percentage points
        (\texttt{required\_sample\_size} in \texttt{soaring.analysis.altitude\_noise},
        with \texttt{proportion\_ci} reporting the point estimate and interval). The
        population is treated as effectively infinite: at $n=2401$ out of $186{,}052$
        flights, the finite-population correction would move $n$ by only $\sim$30, not
        worth the extra parameter. The hang-glider archive (6716 files) stays a full
        census (cheap enough, under two minutes). Total runtime: $\sim$50\,s. Result:
        paragliders $(28.4\pm1.8)\,\%$ (95\% CI, sampled), hang gliders $17.5\,\%$ (exact,
        census).
  \item[Also caught mid-flight] The IGC parser accepted syntactically well-formed but
        semantically invalid \texttt{B} records (e.g.\ minute $\geq 60$, hour $\geq 24$,
        out-of-range latitude/longitude) without rejecting them; added explicit
        format-validity checks (9 new tests) and corrected the thesis wording, which had
        conflated this \emph{implemented} format-validity layer with the fix-level
        \emph{dynamics-plausibility} layer of Table~4.1 -- at the time designed (a
        dataclass of thresholds) but not yet implemented as an enforced filter.
\end{description}

\paragraph{2026-07 --- Pre-processing made data-driven and config-driven.}
Reworked Chapter~4 so the flight-level filtering is computed from the \emph{full parsed
dataset} (paragliders and hang gliders), not the declared catalog metadata.
\begin{description}
  \item[What changed] A track scan (\texttt{soaring.analysis.preprocessing.scan\_tracks})
        parses every downloaded track; Figure~4.2 shows the real duration and
        flown-path-length distributions with per-variable \emph{marginal} retention
        curves. Adopted cuts: duration $\geq 40$ min and path length $\geq 30$ km (a
        minimal cut, $\sim$95\% retained); the earlier min-fix-count cut was dropped as
        redundant with duration, and the \emph{extent} proxy was replaced by total flown
        \emph{path length} (median 86 km para, 158 km hang).
  \item[Key decision] Every pre-processing threshold now lives in
        \texttt{configs/preprocessing.yaml} (never hard-coded), loaded via
        \texttt{load\_preproc\_config} into typed dataclasses.
  \item[Also added] Great-circle (haversine) inter-fix speed/distance \emph{before} the
        ENU conversion; a Savitzky--Golay smoothing/differentiation section with a
        noise-matched procedure for its two hyperparameters; the switch from a common
        resampling cadence to native per-flight $\Delta t$ (Figure~4.5); and an
        implementation blueprint (\texttt{docs/guide/preprocessing-pipeline.md})
        recording the exact schema and steps, to retire into code once the pipeline is
        built.
\end{description}

\paragraph{2026-07 --- Gap diagnostics; scan cache; baro-presence consolidated.}
Added Figure~4.5, the sampling-regularity diagnostics (largest gap relative to native
$\Delta t$, missing fraction of a uniform grid, each with its marginal retention curve),
fulfilling a promise Sec.~4.1.6 already made but had not yet delivered.
\begin{description}
  \item[Scan cache] A fresh full parse would have repeated the $\sim$15--20\,min
        paraglider census just done for Figure~4.2, so the per-flight scan is now
        \textbf{cached} to \texttt{data/<discipline>/track\_scan.parquet} (gitignored,
        like \texttt{catalog.csv}; \texttt{load\_or\_scan\_tracks} reads it if present,
        else scans and writes it --- later moved to the SSD, see below).
        \texttt{track\_stats} was extended with a few near-free byproducts of the same
        scan --- barometric-presence fraction, max horizontal/vertical speed, barometric
        altitude range --- for future validation of Table~4.1's fix-level bounds.
  \item[Consolidation] The barometric-presence fraction (Sec.~4.1.1) now reads from this
        same cache (\texttt{altitude\_noise.baro\_presence\_from\_scan}) instead of
        running its own separate scan, turning the paraglider number from a sized sample
        into an exact census at no extra cost: $30.1\,\%$ (exact, $n=186{,}025$) ---
        consistent with the earlier sampled estimate, $(28.4\pm1.8)\,\%$.
  \item[Also caught] A pre-existing fragility in \texttt{altitude\_noise.collect} where a
        single unreadable file (a transient external-disk hiccup) aborted the whole PSD
        sample; it now skips and logs the file instead.
\end{description}

\paragraph{2026-07 --- SSD reorganised; no data kept locally; figure fixes.}
\begin{description}
  \item[Policy going forward] Raw data, catalogs and every analysis byproduct live only on
        the external SSD, never in the local repo.
  \item[Reorganisation] Each discipline's \texttt{data\_root} is now laid out by maturity
        --- \texttt{raw/\{igc,raw\_xml\}} (untouched acquisition output),
        \texttt{catalog/\{catalog.csv,seasons\_index.csv\}} (tables derived from it),
        \texttt{derived/} (further byproducts, today just \texttt{track\_scan.parquet};
        future cleaned/filtered data and analysis results get their own
        \texttt{processed/}/\texttt{analysis/} siblings), with \texttt{Config} updated
        accordingly (new \texttt{derived\_dir} property). Verified file counts
        before/after the move (paraglider: 186{,}052 tracks, 27 XMLs; hang glider:
        6{,}716 tracks, 25 XMLs --- all matched) and that both acquisition CLIs still
        work against the new layout. Previously-local \texttt{track\_scan.parquet}
        copies moved to \texttt{<data\_root>/derived/}; orphaned local
        \texttt{catalog.csv} copies (superseded by the track-based scan months ago) were
        removed.
  \item[Figure~4.5 (gaps)] The adopted \texttt{max\_gap\_factor} was raised from 5 to 10
        --- at 5 it retained only 77.8\,\%/65.0\,\% of para/hang flights (nowhere near
        the near-full retention of the duration/path cuts), at 10 it reaches
        86.9\,\%/85.1\,\%. Considered 20 (90.8\,\%/90.6\,\%, little further gain) and
        rejected it: the cut is \emph{relative} to each flight's own native $\Delta t$,
        so for the slower hang-glider loggers (up to 10 s native) a factor of 20 would
        tolerate interpolating across a $\sim$200\,s gap --- long enough to span a
        thermal climb. Also tightened the distribution panels' axis range to the data's
        own percentiles (the old 0--200/0--0.6 range was mostly empty).
  \item[Figure~4.1 (altitude noise)] Panel (a)'s difference trace now uses a colour
        distinct from both channels (was reusing GNSS's red); its window grew from 10 to
        30 minutes; a Nyquist-frequency line was added to panel (b) (previously
        discussed in the caption but never drawn). Also found and fixed a real bug in
        the ``representative flight'' selection: picking the \emph{longest flight by
        elapsed time} broke on a 30-hour, 1{,}100-fix outlier (one corrupted timestamp
        gap) --- fixed to select by \emph{fix count} instead, which has no such failure
        mode.
  \item[Local data folder clarified] The two versioned \texttt{seasons\_index.csv} are
        the offline reproducibility anchor for the thesis's dataset counts (the document
        builds with no SSD mounted), not stray local data. Their sync from the canonical
        SSD copy is now explicit and automatic (\texttt{refresh\_seasons\_index.py}, run
        best-effort by the pre-commit hook), so the snapshot cannot silently drift.
\end{description}

\paragraph{2026-07 --- Fix-level cleaning made data-driven; parser roll-over bug; figure and data-folder cleanups.}
\begin{description}
  \item[Fix-level cleaning audited] Table~4.1 is now audited on the data, like the
        flight-level cuts: a new figure in Sec.~4.1.2 shows the per-\emph{fix}
        distributions of horizontal speed, barometric vertical speed and barometric
        altitude (\texttt{make\_fixlevel\_diagnostics\_figure}, from
        \texttt{fix\_level\_distributions} over a seeded sample --- millions of fixes).
        What justifies a fix-level bound is the fraction of \emph{fixes} it removes, not
        the number of flights it touches; the per-flight \emph{maxima} already in the
        scan (\texttt{max\_vxy\_mps} and friends) answer the wrong question, so they do
        not place the cuts.
  \item[Gap handling unified] Removed the absolute 60 s inter-fix time-gap from the
        fix-level table: it duplicated the sampling-regularity cut of Sec.~4.1.6, which
        is more principled (relative to each flight's own native $\Delta t$). Neither was
        ever an enforced filter, so this is a spec simplification, not a behaviour
        change.
  \item[Parser bug fixed] Found the root cause of the ``30-hour, 1{,}100-fix'' pathology
        worked around previously: the IGC parser's midnight roll-over unwrap treated
        \emph{any} backward step in time-of-day as a new day and added 86{,}400 s, so a
        few-second GPS glitch became a spurious multi-hour gap. Now only a drop of
        (nearly) a whole day counts as a roll-over, with residual backward jitter
        clamped to keep the series monotonic. Added regression tests for both the true
        roll-over and the glitch.
  \item[Figure and config cleanups] The altitude-noise figure (Sec.~4.1.1) is now four
        panels --- (a) raw channels, (b) their difference, (c) PSD, (d) fallback rate ---
        fixing the old a/b/d labelling. The gap-diagnostics figure (Sec.~4.1.6) no longer
        clips its tail into an artificial edge spike, with an x-range fitted to the
        data's own percentiles. Result-specific numbers (retention percentages, one-off
        anecdotes) were pruned from YAML/code comments --- such numbers belong in the
        thesis, not in reusable config.
  \item[Local data folder clarified] Rewrote \texttt{data/README.md} and fixed several
        stale references to a flat \texttt{data/seasons\_index.csv} path that never
        existed.
\end{description}

\paragraph{2026-07 --- Coordinate/geodesy figures.}
Redrew the coordinate-transformation figures: Figure~4.3 (WGS84 ellipsoid; geodetic vs
geocentric latitude) and Figure~4.4 (the ECEF$\to$ENU rotation, in our notation), and added
the ESA Navipedia citation for the ECEF conversion.

\paragraph{2026-07 --- Altitude-noise figure (Fig.~4.1): panels (a)/(b) redesigned.}
A closer look at the rendered figure (not caught by any test, since nothing there errors)
found two compounding problems in the one-flight time-series panels, both traced to a
single shared 30-minute window: the several-hundred-metre climb dominated the vertical
scale, hiding the GNSS jitter the panel exists to show; and panel~(b), subtracting only the
scalar \emph{mean} of the GNSS$-$barometric difference, ended up on a $\pm20$\,m scale that
looked inconsistent with the $\sim$130\,m gap plainly visible in panel~(a).
\begin{description}
  \item[First attempt (reverted)] Shortened the shared window to 2 minutes and switched
        panel~(b) to a \emph{linear} detrend (matching \texttt{\_welch}'s). This resolved
        both symptoms but destroyed something worth keeping: the GNSS$-$barometric
        difference actually drifts by tens of metres over a full flight (plausibly the
        barometric channel's departure from the ICAO standard atmosphere as altitude
        changes) --- a real, physically meaningful feature that a 2-minute window is too
        short to show and a linear detrend actively removes.
  \item[Final design] Decoupled the two panels' windows and mechanisms instead of forcing
        one compromise on both. Panel~(a) keeps a short (2-minute) window but now plots
        barometric and GNSS on \emph{two independent $y$-axes} (same numeric span,
        different baseline) rather than one shared axis, which previously had to stretch
        from one channel's minimum to the other's maximum --- wasting on blank space
        exactly the room needed to compare the two channels' roughness; with matched
        spans the GNSS trace is visibly more jagged than the barometric one. Panel~(b)
        reverts to the original 30-minute window with only the \emph{mean} removed (no
        detrend), so the slow drift is visible again alongside the residual jitter, with
        the removed mean offset (139\,m for this flight) annotated directly on the
        panel. Both panels' wording was generalised from ``climb'' to ``vertical
        motion'', since a short window can just as well land on a glide.
\end{description}

\paragraph{2026-07 --- \texttt{dev}/\texttt{analysis} moved from pip extras to uv default dependency groups.}
Chasing the figure fix above, a bare \texttt{uv sync --reinstall-package soaring} (no
\texttt{--extra} flags) silently reset the environment to the core-only dependency set,
uninstalling \texttt{pytest}/\texttt{ruff}/\texttt{mypy}/\texttt{matplotlib}/
\texttt{scipy}/\texttt{pyarrow} --- and, on the way, hit a pre-existing \texttt{uv} bug
(uninstall failing on packages whose \texttt{RECORD} file was missing) that required
rebuilding \texttt{.venv} from scratch.
\begin{description}
  \item[Root cause] \texttt{dev} and \texttt{analysis} were plain PEP~621
        \texttt{optional-dependencies}, so \emph{any} \texttt{uv sync}/\texttt{uv run}
        invocation without matching \texttt{--extra} flags resyncs the venv down to
        exactly what was requested, silently dropping whichever extras the previous
        invocation had added --- alternating flagged and unflagged commands across a
        session means constant install/uninstall churn.
  \item[Fix] Moved both to \texttt{[dependency-groups]} (PEP~735) with
        \texttt{[tool.uv] default-groups = ["dev", "analysis"]}, so every
        \texttt{uv sync}/\texttt{uv run} includes them unconditionally. Neither group was
        ever meant to be pip-installed by an external consumer in practice (this repo is
        not published), so the loss of \texttt{pip install .[dev]}-style installability
        is immaterial; \texttt{docs} remains a real extra since it is genuinely
        on-demand. The docs-deploy CI workflow now opts out of the new default groups
        (\texttt{--no-default-groups}) to stay minimal.
  \item[Docs updated] \texttt{README.md}, \texttt{docs/guide/installation.md}
        (dependency-groups table, pip fallback instructions), the two reporting
        scripts' now-unnecessary \texttt{uv run --with matplotlib --with scipy}
        incantations, and the \texttt{soaring.analysis} package docstring.
\end{description}

\paragraph{2026-07 --- Logbook readability; Next-steps drift; docs deploy on demand.}
Three unrelated loose ends, tidied together.
\begin{description}
  \item[Logbook typography] The narrative notes below were dense, single-paragraph
        entries --- hard to scan. Reformatted the longer ones into short lead sentences
        plus labelled sub-points (\emph{what changed}, \emph{why}, \emph{gotchas}), and
        dropped a stale forward-reference in the altitude-channel entry (``to confirm on
        a larger sample once the SSD is remounted'', superseded two entries later by an
        exact 186{,}025-flight census).
  \item[Next-steps drift] The \nameref{sec:next-steps} list below had fallen out of step
        with the thesis chapter it is meant to mirror: \emph{Preliminary characterization}
        (thesis Sec.~4.2) and \emph{Tooling} (Sec.~4.9) were missing entirely, and the
        whole-trajectory, phase-resolved and transport-analysis bullets were missing
        content the thesis already has (the propagator/return-probability/first-passage
        observables; the step--waiting coupling and phase-sequence memory; the
        stratification and bias-control plan). Re-synced against the current chapter.
  \item[Docs deploy] The GitHub Pages workflow redeployed on every push to \texttt{main},
        including pushes with nothing to do with the site (e.g.\ acquisition/analysis
        code with no docstring change) --- not wanted. One such deploy had also failed
        with a generic, transient ``try again later'' from GitHub's Pages API, but that
        was an isolated glitch, not a config problem (confirmed via the Actions log: only
        that one run failed out of the last ten, and only at the deploy step, after a
        successful build). Fix: the \texttt{push} trigger is now scoped with
        \texttt{paths} to what actually feeds the built site (\texttt{docs/**},
        \texttt{mkdocs.yml}, \texttt{src/soaring/**}, whose docstrings mkdocstrings
        renders into the API reference), so the site still updates itself automatically
        but only for changes that matter to it; \texttt{workflow\_dispatch} stays
        available for an on-demand redeploy.
\end{description}

\paragraph{2026-07 --- License added; README badges; a real test-running CI.}
\texttt{pyproject.toml} had declared \texttt{license = MIT} since the project's start, but
no \texttt{LICENSE} file existed, and the repo had no workflow that actually ran the test
suite (only the docs-deploy one).
\begin{description}
  \item[License] Added \texttt{LICENSE} (MIT): permissive, minimal text, the de facto
        standard for research Python code, and consistent with what \texttt{pyproject.toml}
        already declared -- no reason to pick something more restrictive for code meant to
        be freely reused.
  \item[New CI: \texttt{tests.yml}] Runs \texttt{uv run pytest} with coverage on every push
        to \texttt{main} and every PR. Needed no \texttt{--extra} flags to get
        \texttt{pytest-cov}/\texttt{matplotlib}/\texttt{scipy}/\texttt{pyarrow} -- a direct
        payoff of making \texttt{dev}/\texttt{analysis} uv default groups earlier. Coverage
        XML uploads to Codecov as a best-effort step (\texttt{fail\_ci\_if\_error: false}):
        a Codecov hiccup, or the repo not being linked there yet, must never redden the
        Tests badge.
  \item[README badges] Added five badges (python version, Tests workflow status, Codecov
        coverage, docs-online link, license) at the top of \texttt{README.md}. The Codecov
        one needs a one-time manual step outside this repo -- linking
        \texttt{matteodisante/soaring-anomalous-transport} on \texttt{codecov.io} (GitHub
        login) and, if it asks for one, adding \texttt{CODECOV\_TOKEN} as a repo secret --
        which cannot be done from here.
\end{description}

\paragraph{2026-07 --- Fig.~4.1/4.2 and Appendix~B: closer reading, several fixes.}
A line-by-line pass over the altitude-noise figure, the fix-level figure, and the PSD
appendix, prompted by direct questions about what they actually show.
\begin{description}
  \item[Fig.~4.1 panel~(a), third look] The dual-axis, short-window version (see the
        entry above) did not read, on the page, as clearly showing GNSS is the noisier
        channel, so it was dropped for a shared axis over the full 30\,min window
        instead -- which showed the offset drift directly, but on request the panel went
        back to a dual axis once more, this time over that same 30\,min window rather
        than a short one. Net result: each channel is independently auto-scaled
        (barometric left, GNSS right), so the two curves track almost on top of each
        other (both are the same $\sim$500\,m climb, just at a different baseline) --
        good for reading each channel's own shape, but it means the raw vertical gap is
        no longer a literal reading of the offset. That number was never really this
        panel's to carry: it lives in panel~(b), and the noise-floor claim in panel~(c),
        regardless of which of these three versions panel~(a) went through.
  \item[Appendix~B (PSD) bug] Two cross-references pointed at ``Figure~4.1(b)'' for the
        PSD, which has been panel~(c) since the four-panel redesign; fixed. Also added
        the actual flight yield of the $\Delta t$-matching step, which the text described
        only qualitatively: of 400 sampled flights per discipline, 212
        paragliders but only 114 hang gliders pass it, because hang gliders split
        across several native rates while paragliders are overwhelmingly
        1-second loggers -- so the hang-glider PSD curve is honestly an average
        over that 1-second minority, not the full hang-glider population, now
        stated as such.
  \item[Fig.~4.2 (fix-level), added] The exact sample (800 flights per discipline,
        6{,}633{,}923/4{,}541{,}511 paraglider/hang-glider fix pairs) was missing from
        the caption; added. Also explained two things a reader flagged as confusing on
        sight: panel~(a) (horizontal speed) shows a knee well below the adopted
        40\,m/s bound ($\sim$23/35\,m/s for
        para/hang) -- read as the transition from ordinary gliding speeds to rarer but
        real high-energy manoeuvres (wind-assisted glides, steep spirals) rather than
        GPS error, hence the bound is deliberately kept above it, not moved down to it,
        so as not to discard genuine dynamics; and panel~(b) (vertical speed)'s
        comb-like histogram is a quantisation artefact (integer-metre barometric
        altitude differenced over a 1-second native step can only return an
        integer m/s), not a plotting bug.
  \item[Table~4.1, added] A one-sentence justification for the $-100$\,m
        lower altitude bound, which read as arbitrary: it is not a claim about true
        terrain elevation, but headroom for the barometric channel's own
        non-QNH-corrected offset (Figure~4.1 measures this directly, $\sim$100\,m
        for the flight shown there), so a genuine sea-level take-off can log
        slightly negative without being cut.
\end{description}

\paragraph{2026-07 --- Fix-level speed bounds made per-discipline, on a much larger sample.}
Chasing the loosened-threshold experiment of the previous entry to a conclusion.
\begin{description}
  \item[Sample size] \texttt{FIXLEVEL\_SAMPLE\_PER\_DISCIPLINE} raised from 800 to
        15{,}000: 15{,}000 paragliders (126{,}723{,}499 fix pairs, $\sim$73\,s) and
        all 6{,}716 hang gliders (a full census at this population size,
        37{,}915{,}334 pairs, $\sim$25\,s) -- under two minutes total, and it resolves
        the sparse tail well enough to distinguish real dynamics from a rare artefact,
        which the 800-flight sample could not.
  \item[The hang-glider bump, confirmed and explained] At this sample size the
        45--90\,m/s feature is unambiguous: not a smoothly decaying tail but a distinct,
        much more slowly-decaying plateau (confirmed with a fine, 2.5\,m/s-binned
        histogram -- the decay rate visibly changes between 37.5 and 42.5\,m/s, from a
        steep real decline to a near-flat plateau lasting to $\sim$90\,m/s). A sustained
        \emph{ground} speed of 160--324\,km/h has no credible airborne explanation for a
        hang glider: read as a logger/GPS artefact population, not rare-but-genuine
        manoeuvres.
  \item[Decision: per-discipline bounds, not one shared value] Paragliders and hang
        gliders have different performance envelopes, confirmed by their different
        tail shapes -- a shared bound is either too loose for one or clips real
        dynamics of the other. \texttt{FixLevelThresholds.max\_horizontal\_speed\_mps}/
        \texttt{max\_vertical\_speed\_mps} are now \texttt{dict[str, float]} keyed by
        discipline (\texttt{configs/preprocessing.yaml}, a nested mapping; a
        \texttt{"sailplanes"} entry can be added later with no code change). Vertical
        speed: 12\,m/s (para) / 14\,m/s (hang).
  \item[Horizontal speed bound, revised again: 40 was still ``slowing'', not
        ``stopped''] First adopted at 25/40\,m/s (para/hang), where the fine-binned
        decay rate first visibly changes. On review, with the panel's $x$-range
        widened (next entry) to actually show the far tail, that point turned out to
        still be on the shoulder of the decline, not the flat artefact itself: the
        bin-to-bin ratio only settles at $\approx\!1$ (a genuine plateau, decay
        actually stopped) around 55\,m/s for hang gliders, and the analogous point for
        paragliders is $\approx\!45$, not 25. The 25/40\,m/s pair was therefore
        cutting into the tail-end of real (if rare) dynamics -- a wind-assisted
        glide, a steep spiral -- rather than sitting cleanly past it. Revised to
        \textbf{45\,m/s (para) / 55\,m/s (hang)}, both now Table~4.1's and the
        config's adopted values.
  \item[Figure and table updated] \texttt{make\_fixlevel\_diagnostics\_figure} now
        draws one dashed cut per discipline, colour-matched to its histogram, with a
        per-discipline removed-fraction annotation (previously one shared cut and
        fraction). Table~4.1 gained a row per discipline for the two speed bounds.
  \item[$-100\,$m altitude bound, empirical check (unchanged, kept for the record)]
        Computed the per-flight mean GNSS$-$barometric offset over 1500 flights per
        discipline (broader than the single Figure~4.1 example): median
        $\approx\!90\,$m, 25th--75th percentile roughly $43$--$135\,$m. This is a
        \emph{whole-flight} average, and the offset itself grows with altitude
        (Sec.~4.1.1), so it likely \emph{overstates} the offset specifically near the
        ground; confirms the order of magnitude of $-100\,$m without pinning it down
        precisely. No error found in Figure~4.2 itself.
  \item[$x$-range widened for (a)/(b)] The panels' range was cropping the view almost
        exactly at the cut (the 99.9th percentile of the pooled sample sits right
        there, by construction), hiding how much -- and what -- the bound leaves out.
        Panels (a)/(b) now extend to each discipline's own 99.99th percentile (computed
        per discipline, not pooled, so hang gliders' heavier tail is not swamped by
        paragliders' larger sample); panel (c) (altitude) keeps the original, narrower
        range on purpose -- its cuts sit far out in an otherwise tight, unimodal
        distribution, and widening it the same way mostly shrinks the informative part
        of the panel.
  \item[Panel (b)'s non-empty valleys, explained and checked] The comb (spikes at
        integer m/s) was expected; that the valleys \emph{between} integers were not
        exactly zero was not, and looked like a bug. It is not: not every flight
        samples at 1\,s (Figure~4.7 -- hang gliders spread across several native
        rates), and differencing over a longer native step returns a finer-grained set
        of values (e.g.\ multiples of 0.2\,m/s at a 5\,s cadence) that fill the gaps
        once every cadence is pooled. Checked directly, split by native rate: among
        1\,s-native fixes, $\sim$99\% land within 0.05\,m/s of an integer (the comb,
        essentially pure); among every other native cadence, only 34--42\% do -- the
        other 58--66\% is exactly what shows up between the teeth.
  \item[Panel (b)'s jagged comb, fixed] Separately from explaining \emph{why} the
        valleys were not empty: the comb itself looked bad -- an arbitrary, fine
        (linspace) grid of bins straddles the integer spikes inconsistently depending
        on where each bin edge happens to fall, which reads as noisy up-down jitter
        rather than the smooth decaying envelope the data actually has. Fix: one bin
        per integer, centred on it (edges at half-integers) rather than an arbitrary
        edge grid -- \texttt{make\_fixlevel\_diagnostics\_figure} now takes an
        \texttt{integer\_aligned} flag, set for panel (b) only (vxy is not similarly
        quantised). The panel is now a clean, monotonically-decaying curve; the
        removed-fraction numbers are unaffected (computed from the raw data, not the
        histogram).
  \item[Vertical speed reverted to one shared bound: 13\,m/s] Unlike horizontal speed,
        paragliders' and hang gliders' $|v_z|$ distributions sit close enough together
        (panel (b)) that a per-discipline split buys nothing; a single 13\,m/s bound
        (between the previous 12/14 split) covers both, with one shared grey dashed
        line and one pooled removed-fraction annotation, mirroring how the altitude
        band already works. \texttt{FixLevelThresholds.max\_vertical\_speed\_mps}
        reverted from \texttt{dict[str, float]} to a plain \texttt{float}; the figure's
        two shared-cut helpers (band and single-cut) were merged into one
        \texttt{\_shared\_cut\_panel} that handles either, so panel (b) still gets the
        \texttt{integer\_aligned} treatment above.
  \item[A LaTeX limit hit, and fixed for good: very long captions] Rebuilding after the
        above changes hit \texttt{! Dimension too large} / ``I can't work with sizes
        bigger than about 19 feet'', deep in \texttt{\textbackslash @makecaption}.
        Bisected the Figure~4.2 caption by half repeatedly to isolate it -- not a typo
        (braces and math-mode \$ were balanced throughout), but the caption's sheer
        length ($\sim$3900 characters): plain \LaTeX{}'s single-line-fit check for
        captions measures the \emph{whole, unwrapped} caption as one hbox before
        deciding whether to wrap it, and past a few thousand characters that
        measurement alone overflows \TeX's $\sim$19-foot maximum dimension, even though
        the wrapped result would render fine. Loading the \texttt{caption} package does
        the same check by default (to centre genuinely short captions), so
        \texttt{singlelinecheck=false} was needed on top of it. This is a general limit
        of this document now that several figures carry long, data-justified
        captions -- not specific to this one -- so the fix (\texttt{main.tex} preamble)
        covers all of them going forward, not just Figure~4.2.
\end{description}

\paragraph{2026-07 --- Figure~4.7 (native sampling interval): discrete data, discrete plot.}
Raised independently: the density in Figure~4.7 rose above 1, and native $\Delta t$ is a
discrete quantity, so why plot a density at all. Checked on the full census: $\Delta t$
lands on an exact whole second for 99.9\% of paragliders and 100\% of hang gliders --
essentially a categorical variable (one of a handful of configured logger rates), not a
continuous one. The previous plot used arbitrary 0.5\,s-wide bins on this discrete data
(\texttt{np.arange(0.25, 8.75, 0.5)}): most paragliders' mass concentrated in the single
bin straddling $\Delta t=1$, and \texttt{density=True} divides a large fraction by that
narrow bin width, which is exactly how it exceeded 1 -- a real number, not a bug, but a
genuinely confusing one for data this concentrated. Fixed the same way as the
vertical-speed comb: bins one full second wide, centred on each integer ($1,\dots,10$,
plus a $\geq\!11$ bin for the long thin tail), so each bar reads directly as ``fraction
of flights at exactly this rate.'' With unit-width bins \texttt{density=True} is
numerically identical to that fraction (dividing by a width of 1 changes nothing), so
the axis is now labelled \emph{fraction of flights}, not \emph{density} -- accurate
either way, but only one of the two labels says what is actually being counted. Real
distribution, now stated precisely in the caption instead of the vaguer ``$1,2,5,\geq\!
8$'' before: paragliders overwhelmingly (69\%) at 1\,s; hang gliders spread across
1\,s (30\%), 5\,s (27\%), 10\,s (15\%), 2\,s (13\%), and a long thin tail beyond 11\,s --
a fuller picture than the old caption's ``up to 8\,s,'' corrected in Sec.~4.1.6 too.

\paragraph{2026-06 --- Documentation.}
Project docs are built with MkDocs and published on GitHub Pages. Initially deployed with
\texttt{mkdocs gh-deploy}, which accumulated one built-site commit per deploy on a
\texttt{gh-pages} branch; migrated to the official GitHub Pages \emph{artifact} workflow
(build $\to$ upload artifact $\to$ deploy), removing the branch and keeping the git history
clean.

\paragraph{Operational note.} In this repo \texttt{uv sync} occasionally fails to
(re)install the editable package; \texttt{uv sync --reinstall-package soaring} fixes it.

\section{Next steps}
\label{sec:next-steps}
These mirror the planning chapter of the state document
(\texttt{thesis/sections/04-next-steps}, itself titled ``Next steps'') and are kept in
sync with it, section by section. The near-term plan is to analyse the data \emph{in
parallel} with studying the segmentation, since many observables can be built before any
phase splitting.

\begin{itemize}
  \item \textbf{Data sources.} FFVL paraglider and hang-glider CFD data are collected,
        cataloged and verified. Extend to further sources -- notably WeGlide for
        sailplanes (a higher-quota API request is pending; the default rate limit makes a
        bulk export infeasible) and possibly XContest -- merging into one catalog and
        stratifying by source. Refresh the per-discipline \texttt{seasons\_index.csv} via
        \texttt{build-catalog} so the document statistics stay current.
  \item \textbf{Pre-processing (thesis Sec.~4.1).} The parser is in place
        (\texttt{soaring.analysis.igc}) and the pipeline is designed, thresholded
        (\texttt{configs/preprocessing.yaml}) and audited on real data: fix-level
        cleaning (great-circle speeds), $v_{xy}$-based trimming, data-driven
        flight-level filtering (duration $+$ path length), the ENU conversion (barometric
        vertical), native-$\Delta t$ uniformity, and Savitzky--Golay
        smoothing/differentiation. Next: implement it end-to-end as production code
        (parser $\to\dots\to$ Parquet \texttt{fixes}/\texttt{flights\_meta} tables) per
        the blueprint, and measure the real ENU power spectra to finalize the two
        Savitzky--Golay smoothing timescales (still placeholders in the config).
  \item \textbf{Preliminary characterization (Sec.~4.2).} Lightweight, segmentation-free
        checks before the main analysis: per-flight trajectory statistics (duration, path
        length, fix count, native $\Delta t$); an isotropy check (the per-component
        variance ratio $\langle E(t)^2\rangle/\langle N(t)^2\rangle$, expected identically
        $1$, plus a Kolmogorov--Smirnov test) needed before the marginal-propagator
        scaling can be used; compatibility across strata (glider class, season) via
        overlaid per-component displacement-variance curves, to justify -- or rule out --
        pooling them; and a spatial-coverage check (take-off and displacement maps) for
        possible geographic stratification.
  \item \textbf{Whole-trajectory observables (Sec.~4.3, no segmentation).} Characterise
        the data with observables that need no phases: the MSD and its scaling exponent
        $\alpha$ (the primary anomalous-transport test); the time-averaged MSD, whose
        agreement or disagreement with the ensemble MSD is an ergodicity test; the
        one-dimensional marginal propagator and its scaling collapse; the return
        probability and the non-Gaussian parameter; the velocity autocorrelation and
        turning-angle distribution; first-passage/first-exit times; the
        speed/vertical-velocity distributions; and geometric measures (tortuosity,
        fractal dimension, radius of gyration), read both as one number per flight and as
        functions of scale. Goal: a first physical/statistical picture, independent of
        any phase splitting.
  \item \textbf{Segmentation (Sec.~4.4).} Study J\'er\'emie's internship report and code,
        then re-examine the transition/search/climb HMM (binary features: sign of
        vertical speed, curvature-angle persistence, straightness) --- decide whether to
        keep that approach or add complexity for a better segmentation on our larger,
        more heterogeneous data.
  \item \textbf{Phase-resolved observables (Sec.~4.5).} After segmentation: the
        waiting-time decomposition (climb $\tau_c$, search $\tau_s$, and their sum
        $\tau_w$), the glide step-length/duration joint law, per-flight anatomy (thermal
        count, phase time/distance budget), and the angular diffusion between successive
        glides. Plus two couplings that decide which transport model applies: the
        step--waiting coupling ($\ell$ vs $\tau_w$, the L\'evy-walk signature) and memory
        in the phase sequence (transition-matrix structure, duration correlations), which
        test the renewal assumption behind the textbook CTRW/L\'evy-walk results.
  \item \textbf{Reproduce the reference (Sec.~4.6).} Check that the enlarged
        paraglider-and-hang-glider dataset reproduces Vilpellet et al.: per-phase
        conditional MSD (their Fig.~3) and per-phase descriptive statistics (their
        Fig.~4), plus the global $H\approx0.88$ --- validation and universality test (the
        sailplane comparison becomes reachable once that source is in hand).
  \item \textbf{Transport analysis (Sec.~4.7).} Estimate the MSD scaling exponent, fit
        the tails of the waiting-time and step-length distributions with principled
        methods (e.g.\ maximum-likelihood power-law fits with goodness-of-fit tests), and
        perform CTRW-vs-L\'evy-walk model selection using the step--waiting coupling.
        Results will be stratified (season, site, glider class, data source) and
        trajectories normalized, with finite-size, censoring and sampling biases
        controlled throughout.
  \item \textbf{Minimal model (Sec.~4.8).} A minimal stochastic model for
        $H\approx0.88$: resolve transition segments, merge search+climb into one
        effective state, with angular diffusion between successive transition
        directions (preliminary idea, to be refined/changed).
  \item \textbf{Tooling (Sec.~4.9).} Largely underway already: the analysis package
        mirrors acquisition, and the dataset statistics plus several diagnostic figures
        (pre-processing, altitude noise) already regenerate automatically from the data
        rather than being pasted in by hand. Remaining: extend the same treatment to
        whatever tables/figures the transport analysis produces, once it exists.
\end{itemize}

\end{document}
